\section{Conclusion and Future Work}
\label{sec:conclusion}

We establish constitutive enforcement at the decoding boundary -- positive security all the way down to the generative act \cite{saltzer1975,anderson1972} -- as a structural control for LLM agent tool calls within the Tantalus agent system (\S\ref{sec:method}), and do so at scale, with a soundness core and an explicit cost axis. The central distinction is \emph{constitutive vs.\ corrective} enforcement: GCD (Condition~\texttt{c}) and a post-parse allowlist (Condition~\texttt{d}) enforce the \emph{same} positive policy over the same language \(L(G_s)\) -- both compute it through the single recognizer \texttt{tantalus\_grammar::allowlist\_verdict}, so the policy is correct-by-construction \cite{sassaman2013langsec} and \(G_s\), compiled from the authenticated principal's identity, is an unforgeable per-request capability \cite{dennisvanhorn1966,miller2006}; they differ only in where that policy sits relative to generation. Under \texttt{c} the policy \emph{is} the output space, so an out-of-scope action is ungenerable -- it never exists as an artifact in the output buffer, logs, telemetry, stream, or KV cache. Under \texttt{d} an unbounded default-allow generator \emph{produces} the malicious action in full and a separate gate then rejects it, re-introducing one layer down the very failure modes positive security exists to escape: the artifact is produced (cost plus an output-channel forensics surface), the gate carries a complete-mediation obligation on every output path \cite{saltzer1975}, and a TOCTOU window opens. GCD's positivity is enforcement-deep -- positive security all the way down to the generative act -- while the allowlist's positivity is policy-deep only; the two are policy-equivalent, with enforcement-completeness differing. Post-generation policy enforcement (Condition~\texttt{d} and systems such as CaMeL \cite{camel2025}, Progent \cite{progent2025}, OAP \cite{oap2026}, and AgentArmor \cite{agentarmor2025}) is policy-deep; the decode-time neighbours Agent-C \cite{agentc2025} and PackMonitor \cite{packmonitor2026} share the enforcement-deep lane but enforce temporal and hallucination constraints respectively, not authorization over enumerable sinks -- see \S\ref{sec:related} for the full distinctions. Five datasets across three pretraining lineages substantiate this. On a deliberately weak workhorse (Qwen3-1.7B-NVFP4), a generator\(\times\)gate factorial at \(\sim\)282{,}000 attack trials per cell reproduces the structural result decisively: the GCD arms emit \emph{zero} out-of-scope artifacts (scored by effect: the attacker URL plus marker must traverse the free-text surfaces that remain open under blind-\texttt{c} -- \texttt{searchFiles.query} and \texttt{respondToUser.message} -- via \texttt{tantalus-pipeline::wins}) and render the gate inert (\texttt{c\_d\_r3} emission \(0\), \texttt{gate\_rej} \(0\)), while behavioral generators behind the identical gate produce 104{,}615--120{,}140 out-of-scope calls each at roughly twice the output tokens. On a capable non-Qwen anchor (Mistral-Small-4-119B-NVFP4), the behavioral arms run \emph{ungated} and deliver what the weak victim could not: RQ1, with the full stack \texttt{a4} still leaking \(1.82\%\) [1.43, 2.29] (CI-lower \(>0\)) and \texttt{a1} (\(56.50\%\)) \(\le\) Control (\(56.77\%\)); and the registered H2 contrast, with \texttt{c} at \(0.00\%\) [0, 0.09] strictly below every behavioral point estimate. GCD also decouples output validity from model capability: on the capable Mistral-119B anchor, behavioral arms produced parseable tool calls only \(\approx59\%\) of the time while \texttt{c} achieved 99.8\% (Tab.~\ref{tab:mistral}), a validity dividend that is independent of bypass rate and holds regardless of attack potency, though its magnitude is victim-dependent: \(\approx41\,\)pp on the Mistral anchor (behavioral \(\approx59.2\%\) vs.\ \texttt{c} 99.8\%) but near-zero on Gemma, whose native free-generation is already 97.9--100\% valid (Tab.~\ref{tab:gemma}). A separate abliterated-adversary steelman -- a maliciously prompted Qwen3.5-27B that is demonstrably willing and able (\(97.85\%\) bypass unconstrained \cite{arditi2024}) -- is driven to \(0.00\%\) by the grammar alone (\(97.85\%\rightarrow0\%\), grammar the sole IV), while provably ingesting its skill's sensitive artifact in \(96.8\%\) of trials. This last result is a rebuttal exhibit, not main statistics: it is engagement evidence (the model engaged the malicious content), not a confidentiality result, and \texttt{c\_sealed}'s \(0\%\) rests on both the structural guarantee (attacker sink ungenerable) and the empirical forced-trajectory trace (96.8\% of trials provably ingested the sensitive artifact yet emitted 0/2{,}000 attacker \texttt{fetchUrl} calls).

The honest envelope is narrower than the headline. External validity rests on three pretraining lineages (Qwen, Mistral, Gemma); the registered \(\ge\)3-lineage bar is met on a point-estimate basis -- three pretraining lineages replicate H1 and H2 (Gemma anchor: control \(98.85\%\), \texttt{c} \(0.00\%\) [0, 0.18], \texttt{a4} \(33.80\%\) [31.73, 35.92] CI-lower \(>0\)) -- but the Gemma anchor is a single-anchor result at \(n=2{,}000\) attack trials/cell (vs.\ 4{,}000 on Mistral and 282{,}000 on the 1.7B) and a single vLLM engine, so full replication-weight closure awaits a fourth lineage (\S\ref{sec:limits}). The RQ1 leak and the formal H2 contrast are carried by the capable Mistral and Gemma anchors, not by the 1.7B (which is sub-floor by design at \(18.33\%\) Control bypass and is the weak-end apparatus, not the headline victim). The registered powered 2pp non-inferiority TOST clustered by \texttt{task\_id} was run on the Mistral anchor (\texttt{mistral119\_ladder.db}, 1{,}456 legit trials/cond, \texttt{--legit-rounds 7}) and confirmed not established: observed \texttt{c}\(-\)Control \(=-1.85\,\)pp (inside the 2\,pp margin) but cluster-bootstrap 95\% CI [\(-7.07\), \(+3.43\)] (half-width \(\approx5.3\,\)pp \(\gg\) 2\,pp), so the CI does not clear the non-inferiority margin even with the powered design (H4 not established). ``0 invalid outputs'' is likewise not literally met on vLLM (regular-\texttt{c} \(98.9\%\) valid, \texttt{c\_closed\_d\_r3} \(99.9\%\), \texttt{c\_guided\_d\_r3} \(97.8\%\), all from free-string truncation under Kleene \texttt{*} that fails safe). The C-guided (L3-closed) superiority result (H6) -- \texttt{c\_l3\_closed} bypass \(0\), deflection \(0\), legit-success \(100.0\%\) -- holds across all three lineages: Qwen-1.7B (624/624), Mistral-119B ladder CP95 [99.75, 100], and Gemma-31B CP95 [99.41, 100]. Grammar-compile time (H5) is measured out-of-band but not met at cold-compile: blind-\texttt{c} p99 \(\approx\) 59--61\,ms (Qwen/Mistral tokenizers) against the registered 25\,ms target; the spirit holds amortized (\(\approx\)0.003\,ms per trial at cache hit). The single subject system (Tantalus) makes this a feasibility claim, not a population-level one; the corpus is selected-on-outcome (the live-malware analogue), so bypass rates are conditional on attack potency and large \(N\) tightens confidence intervals \cite{clopper1934} without buying generalization. The class-level ``eliminates'' claim is carried by the soundness theorem, which quantifies over all logit distributions and so holds against a fully injection-compromised model, not by any sample; proof and evidence stay in separate lanes.

\paragraph{Future work.} The open items follow directly from this envelope:
\begin{itemize}[leftmargin=1.4em]
  \item \textbf{A fourth pretraining lineage} (Llama-family or a dense Gemma variant) to replicate H1/H2/H3/H6 beyond the current three (\S\ref{sec:limits}); \texttt{Nemotron-3-Super-120B} was disqualified due to hybrid-mamba decode throughput (\(\approx\)64--72\,tok/s aggregate, \(\approx\)30\(\times\) too slow for a full ladder) despite passing all potency gates.
  \item \textbf{Capable-anchor C-guided and C+ cells.} The Mistral run used blind \texttt{c} only; running \texttt{c\_guided} and \texttt{c\_plus} on a capable model would extend the H6 superiority result and the closed-response validity result beyond the weak 1.7B.
  \item \textbf{Tightening the H4 non-inferiority CI.} The powered 2pp TOST clustered by \texttt{task\_id} (\texttt{--legit-rounds 7}, 1{,}456 trials/cond) has been run and fails -- the cluster-bootstrap half-width (\(\approx\)5.3\,pp) exceeds the 2\,pp margin -- so the open item is to explain why the cluster-level precision is insufficient and what sample size or cluster structure (more \texttt{task\_id} clusters, more rounds per cluster) would shrink the half-width below the margin.
  \item \textbf{Closing the cold-compile H5 deviation.} Cold FSM compile (xgrammar \cite{dong2024xgrammar} / guidance) is measured at p99 \(\approx\)59--61\,ms against the registered 25\,ms target; closure on a production-scale tokenizer or a tighter grammar is outstanding (the amortized per-trial cost \(\approx\)0.003\,ms already meets the spirit at cache hit).
  \item \textbf{The full mechanized soundness proof} as a companion artifact, promoting the proof sketch's caveats to discharged hypotheses and making the for-all-logit lemma and the tokenizer-lifting argument first-class.
  \item \textbf{Paper-2: free-text confidentiality} -- enumerable-response \texttt{c\_plus} where the response space is enumerable, plus layered DLP for genuinely open-vocabulary channels (\texttt{searchFiles.query}, composed \texttt{respondToUser.message}) that lie outside this paper's action-integrity scope.
\end{itemize}

The thesis, restated, is structural and narrow: enforcement at the decoding boundary -- the grammar mask applied at the sampler before any token is drawn -- is positive security all the way down to the generative act \cite{saltzer1975,anderson1972}. The transformer's logits encode intent and may be arbitrarily compromised; the generated tokens are constrained regardless, so the unauthorized action is not caught but ungenerable, with emitted \(\in L(G_s) \subseteq \mathrm{Authorized}(s)\).
