\makeatletter
\@ifundefined{c@corollary}{\newtheorem{corollary}{Corollary}}{}
\makeatother

\section{Formal Core: A Soundness Sketch}
\label{sec:formal}

Host-security benchmarks have established an empirical-measurement paradigm \cite{sassaman2013langsec,zhang2024asb}. An empirical block rate, however, reports only that some fraction of attacks was stopped: a finite measurement that licenses no statement about the next, unseen sample. Beyond that paradigm, this work adds a cost/economics axis and, the subject of this section, a \emph{soundness theorem}. Grammar-constrained decoding admits a class-level guarantee: under the conditions stated below, an emitted action provably lies within the authorized set, \emph{independent of the model's logits} and therefore against a fully injection-compromised model. This section states that guarantee precisely and is scrupulous about its scope and its trusted computing base (TCB). We present it as a \emph{sketch}: the fully detailed proof, with the tokenizer-lifting argument made first-class, is the companion artifact \texttt{docs/proof.md} (available at \url{https://github.com/cybersharkvin/gcd-authz/blob/main/docs/proof.md}). We do \emph{not} claim a complete mechanized proof here.

\subsection{Setup}
\label{sub:formal-setup}

Fix a tokenizer with finite vocabulary \(V\). A generation is a finite sequence of tokens \(t_1 t_2 \cdots t_n \in V^{*}\). For an authenticated principal with scope \(s\), the \(G_s\) compiler (the novel artifact, Section~\ref{sec:method}) turns \(s\) into a context-free grammar \(G_s\) whose terminals are tokenizer tokens; authorized accounts, endpoints, channels, and paths become ground terminals, and out-of-scope parameters are absent from the language \(L(G_s) \subseteq V^{*}\). We write \(\mathrm{Authorized}(s) \subseteq V^{*}\) for the set of token sequences whose decoded action is permitted for principal \(s\).

This is the LANGSEC programme's correct-by-construction recognizer principle applied to the token sampler \cite{sassaman2013langsec}: the accepted language \emph{is} the policy, and anything outside it is not merely rejected but ungenerable. The compiled \(G_s\) acts as an unforgeable per-request capability over the action space \cite{dennisvanhorn1966,miller2006}. It cannot be forged by an injected instruction, because it is compiled from the authenticated identity and applied at the decoder, outside the attacker-reachable input channel \cite{dennisvanhorn1966,miller2006}.

The constrained decoder applies, at each decode step \(i\), a \emph{grammar mask} computed from the prefix \(t_1 \cdots t_{i-1}\) already emitted. The mask drives to zero the probability of every token that cannot extend that prefix to a valid prefix of some string in \(L(G_s)\), \emph{before} sampling. Formally, let \(\mathrm{Ext}_{G_s}(p) \subseteq V\) denote the set of tokens \(t\) such that \(p\,t\) is a prefix of some member of \(L(G_s)\) (with a distinguished end-of-sequence token \(\mathrm{EOS}\) that is disjoint from \(V\) and governed by a separate admission rule: \(\mathrm{EOS}\) is admissible at step \(i\) iff \(t_1 \cdots t_{i-1} \in L(G_s)\), matching the treatment in the companion artifact). The decoder restricts the support of the next-token distribution at step \(i\) to \(\mathrm{Ext}_{G_s}(t_1 \cdots t_{i-1})\), with the end-of-sequence token permitted exactly when the prefix is itself in \(L(G_s)\). This is the standard push-down-automaton mask realized by xgrammar \cite{dong2024xgrammar} and llama.cpp GBNF \cite{gerganov_llamacpp}, in the same lineage as guidance-style constrained generation \cite{willard2023}. Enforcement happens at the \emph{sampler}, not the model: the transformer always computes a full logit distribution over \(V\), and under injection it may assign an arbitrarily high logit to a forbidden token; the mask is applied to that distribution prior to any sampling decision.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, align=center, minimum height=9mm, text width=20mm, font=\scriptsize, inner sep=2pt},
  arr/.style={->, thick}
]
\node[box] (lm) at (0,0) {Transformer};
\node[box] (logits) at (2.6,0) {full logit distribution over \(V\)};
\node[box, fill=black!8] (mask) at (5.2,0) {grammar mask \(\mathrm{Ext}_{G_s}\)(prefix)};
\node[box] (renorm) at (7.8,0) {renormalized: forbidden tokens at prob.\ \(0\)};
\node[box] (samp) at (10.4,0) {sampler (greedy / top-\(k\) / top-\(p\) / temp.)};
\node[box] (out) at (13.0,0) {token \(\in \mathrm{Ext}_{G_s}\)(prefix)};
\draw[arr] (lm) -- (logits);
\draw[arr] (logits) -- (mask);
\draw[arr] (mask) -- (renorm);
\draw[arr] (renorm) -- (samp);
\draw[arr] (samp) -- (out);
\draw[arr, dashed] (out.south) to[out=-90,in=-90,looseness=0.4] node[below, font=\scriptsize] {prefix \(\gets\) prefix \(+\) token (next step)} (lm.south);
\end{tikzpicture}
\caption{Where enforcement happens: the grammar mask is applied to the full logit distribution \emph{before} sampling, at every step, independent of the logit values. A forbidden token assigned high logit by the model under injection is zeroed out and cannot be sampled. The diagram covers the token-level guarantee; an additional harness-level parse step (not shown) converts the emitted token sequence to a structured \texttt{ToolParams}. Free-text truncation at \texttt{max\_tokens} produces a token sequence that is grammar-legal but harness-unparseable (it fails safe; Section~\ref{sec:limits}), which is why the empirical validity rates sit just below \(100\%\) (Corollary~\ref{cor:validity}).}
\label{fig:mask-locus}
\end{figure}

\subsection{Mask Soundness}
\label{sub:formal-lemma}

\begin{lemma}[Mask soundness]
\label{lem:mask-soundness}
Fix \(G_s\) and run the constrained decoder above. For \emph{any} sequence of logit distributions over \(V\) produced by the model and \emph{any} sampler --- greedy \(\arg\max\), top-\(k\), top-\(p\) (nucleus), or temperature sampling at any temperature --- every token drawn at step \(i\) lies in \(\mathrm{Ext}_{G_s}(t_1 \cdots t_{i-1})\). Consequently, any generation the decoder reports as complete is a member of \(L(G_s)\).
\end{lemma}

\begin{proof}[Proof sketch]
At step \(i\) the mask sets the probability of every token outside \(\mathrm{Ext}_{G_s}(t_1 \cdots t_{i-1})\) to zero before sampling. Each of the named samplers selects a token from the support of the (renormalized) distribution it is given: \(\arg\max\) returns a token of maximal probability, which has nonzero probability; top-\(k\), top-\(p\), and temperature reweighting all preserve the zero set --- a token assigned probability zero remains unselectable under any of them. Hence the drawn token has nonzero post-mask probability and therefore lies in \(\mathrm{Ext}_{G_s}(\cdot)\). This holds at every step regardless of what the logits were, since the argument never inspects the logit values, only the support after masking. By induction over the \(n\) steps, the prefix invariant ``\(t_1 \cdots t_i\) is a valid prefix of \(L(G_s)\)'' is maintained, and completion is admitted only when the prefix is in \(L(G_s)\); therefore the completed generation is a member of \(L(G_s)\). (The induction reaches a completed generation only if the grammar satisfies the progress condition --- for every reachable valid prefix either a legal extension exists or \(\mathrm{EOS}\) is admissible; this grammar-construction discipline is stated and discharged in the companion artifact \texttt{docs/proof.md} and is guarded in the implementation by the structural loop-terminator and the closure discipline on free-text fields.) A complete treatment, including the end-of-sequence condition and the tokenizer lift of Assumption~\ref{asm:tokenizer-lift}, is deferred to \texttt{docs/proof.md}. The underlying PDA-mask mechanism is due to \cite{willard2023,dong2024xgrammar}.
\end{proof}

The clause ``for any logit distribution'' carries the argument: the lemma assumes nothing about the logits, so it covers the adversarial case in which a prompt injection has driven the model to \emph{intend} a forbidden action. Logits encode intent and are uncontrolled. The generated tokens, however, are constrained.

\begin{remark}[Relation to other decode-time enforcement]
\label{rem:nearest-neighbor}
Two recent decode-time enforcement systems act at the same point in decoding, and they separate from this work along two distinct axes: (a) whether the mask is applied \emph{before} sampling so that the forbidden token never carries positive probability and no post-sampling check or resample is needed (Figure~\ref{fig:mask-locus}); and (b) whether the authorized language is grounded in the authenticated principal identity, compiled per request, rather than in a fixed domain-specific constraint. Agent-C \cite{agentc2025} proves temporal-ordering soundness for agent steps via an SMT check-and-resample mechanism: criterion (a) separates it from us, since its enforcement runs after a candidate is produced. PackMonitor \cite{packmonitor2026} eliminates package-name hallucinations via a DFA-driven logit mask, a mask-before-sampling architecture structurally similar to ours, so criterion (a) does \emph{not} separate it. The separation is criterion (b): PackMonitor constrains package-name enumeration for hallucination-correctness under a fixed constraint domain, whereas \(G_s\) is a per-request authorization grammar compiled from authenticated caller identity. We claim to be the first to build and measure generation-time \emph{authorization} at this scale, not the first to conceive decode-time enforcement (Agent-C and PackMonitor, above).
\end{remark}

\begin{corollary}[Validity as a mechanism property, not a capability property]
\label{cor:validity}
Under Lemma~\ref{lem:mask-soundness}, every emitted token is grammar-legal and every \emph{completed} generation (given the progress discipline of the companion artifact) is a well-formed member of \(L(G_s)\), independent of the logits or model size. In particular, a smaller model constrained by \(G_s\) achieves the same structural validity guarantee as a frontier model constrained by \(G_s\).
\end{corollary}

\noindent
This makes GCD's tool-call validity a property of the \emph{enforcement mechanism}, not of model size or competence. The same Lemma~\ref{lem:mask-soundness} yields both halves of the contribution from one mask: an out-of-scope token is ungenerable (security), and only well-formed in-grammar tokens are generable (validity). The empirical consequence is quantitative and surprising. As an empirical illustration (Section~\ref{sec:results}), the 1.7B generator under GCD emits a well-formed in-policy tool call \(98.9\%\) of the time (\texttt{c\_d\_r3}; ledger \S1, Table~\ref{tab:blast}), while the capable 119B Mistral anchor in unconstrained free generation does so only \(59.2\%\) of the time (\texttt{control}; Table~\ref{tab:mistral}); the same Mistral under GCD reaches \(99.8\%\) (\texttt{c}; Table~\ref{tab:mistral}). The smaller model with the grammar beats the larger model without it on structural validity. The two axes are separable in deployment: capability buys task intelligence, the grammar buys structural validity, so a cheaper model under GCD gets structurally guaranteed valid calls that a frontier model fails to produce roughly \(40\%\) of the time unaided. The residual invalidity in GCD cells (\(0.1\)--\(2.2\%\); worst case \texttt{c\_guided\_d\_r3} at \(97.8\%\), Section~\ref{sec:res-blast}, Table~\ref{tab:blast}) is free-text truncation at \texttt{max\_tokens}. A truncated generation is \emph{not} a grammar nonconformance, since every emitted token was grammar-legal, but it is unparseable, and it fails safe (outcome \texttt{no\_action}, never a bypass; Section~\ref{sec:limits}). Corollary~\ref{cor:validity} covers all \emph{emitted tokens}; truncation-invalids are a harness-level parse failure, not a soundness failure. \emph{Fence:} the grammar guarantees the call is valid and in-policy. It does not guarantee the call is \emph{correct} about which in-scope action the task requires: that remains a function of model intelligence and task grounding (the deflection DV). The mechanism never fails \emph{structurally} by construction; it can still choose the wrong in-scope action.

\subsection{The Soundness Theorem}
\label{sub:formal-theorem}

\begin{theorem}[Action-integrity soundness]
\label{thm:soundness}
Let \(\textit{emitted}\) be any completed generation of the constrained decoder for principal scope \(s\), and suppose the compiled grammar satisfies \(L(G_s) \subseteq \mathrm{Authorized}(s)\) (Assumption~\ref{asm:tcb}). Then
\begin{equation}
\label{eq:soundness}
\textit{emitted} \in L(G_s) \subseteq \mathrm{Authorized}(s),
\end{equation}
so the emitted action is within the authorized set --- independent of the logits, hence against a fully injection-compromised model.
\end{theorem}

\begin{proof}[Proof sketch]
By Lemma~\ref{lem:mask-soundness}, \(\textit{emitted} \in L(G_s)\) for any logits and any sampler. Assumption~\ref{asm:tokenizer-lift} ensures that token-level membership in \(L(G_s)\) implies character-level authorization (the decoded action \(\delta(\textit{emitted})\) lies in the intended authorized set); Assumption~\ref{asm:tcb} then gives \(L(G_s) \subseteq \mathrm{Authorized}(s)\) in the token-lifted sense. Compose the inclusions to obtain \eqref{eq:soundness}.
\end{proof}

\paragraph{What this licenses, and what it does not.} Theorem~\ref{thm:soundness} is a statement about \emph{every} producible output of the mechanism, not about a measured fraction of them. It is therefore the one asset that supports a class-level \emph{eliminates-by-construction} claim (out-of-scope actions are ungenerable) that no finite sample could establish however large. The empirical samples reported in Section~\ref{sec:results} do \emph{not} carry the ``eliminates'' claim: they include \(\ge\!868{,}000\) GCD attack trials at \(0\) bypass (Section~\ref{sec:results}, Table~\ref{tab:ledger}), of which the \(2{,}000\) steelman trials are partly true-by-construction. What they show is that behavioral controls genuinely leak at scale (Section~\ref{sec:res-mistral}) and that the implementation matches the theory. This is the distinction between evidence and proof that a correct-by-construction recognizer \cite{sassaman2013langsec} licenses. Proof and evidence stay in separate lanes.

\subsection{Trusted Computing Base and Caveats}
\label{sub:formal-tcb}

The guarantee is conditional, and we state the conditions honestly.

\begin{assumption}[Compiler / allowlist correctness --- the TCB]
\label{asm:tcb}
The \(G_s\) compiler is trusted: Theorem~\ref{thm:soundness} holds \emph{iff} the compiled language satisfies \(L(G_s) \subseteq \mathrm{Authorized}(s)\). If the allowlist is wrong --- it admits a string whose decoded action is not in fact permitted --- the theorem's conclusion follows from a false premise and provides no protection.
\end{assumption}

The compiler is therefore the trusted computing base, exactly as the policy engine is the TCB in any positive-security control. We mitigate this dependency, but do not eliminate it, with two independent checks: (a) an automated \emph{acceptance oracle} that validates each compiled grammar against the target engine before any run (and revalidates per engine; Section~\ref{sec:method}); and (b) an \emph{executor re-validation backstop} that re-checks the decoded action against the same allowlist at the point of execution. The backstop is itself a Condition-D-class corrective check (the same \texttt{tantalus\_grammar::allowlist\_verdict} routine both \(C\) and \(D\) call; Section~\ref{sec:method}), deployed here as defense-in-depth that \emph{complements} the constitutive control rather than contradicting it: the same post-parse check is the recommended corrective backstop for \(C\) (Section~\ref{sec:discussion}). C and D are policy-equivalent over \(L(G_s)\); they differ in enforcement completeness, and using one to guard the TCB of the other is consistent with that framing.

\begin{assumption}[Sound tokenizer lift]
\label{asm:tokenizer-lift}
The grammar \(G_s\) is authored at the character level (URLs, paths, channel identifiers), but the mask of Section~\ref{sub:formal-setup} operates over the \emph{token} language. Correctness requires a sound lift from the character-level grammar to the tokenizer token language --- the token-level language must decode to exactly the intended character-level language, with no over- or under-approximation introduced by token boundaries.
\end{assumption}

The tokenizer lift is an engineering obligation, not a free assumption: a faulty lift can admit a token sequence that decodes to an out-of-scope string, silently violating Assumption~\ref{asm:tcb} one layer down. We discharge it operationally, per engine, via the acceptance oracle, which asserts that the engine actually constrains output to the intended language on probe inputs, catching silent-ignore and tokenizer-misalignment footguns. We make it first-class in the companion proof.

\begin{remark}[Scope]
\label{rem:formal-scope}
Theorem~\ref{thm:soundness} covers \emph{action integrity over enumerable-sink tools}: structured tool calls whose authorized parameter values form an enumerable language (endpoints, channels, paths, message identifiers). It does \emph{not} cover free-text confidentiality through parameters that must remain unconstrained, such as \texttt{searchFiles.query} or \texttt{respondToUser.message}. A free-text field is, by construction, not narrowed by \(G_s\), so the theorem makes no claim about what flows through it. Confidentiality through unconstrained free-text channels is Paper-2 scope, addressed there by complementary measures (enumerable-response closure for the enumerable-response agent class, and layered detection). The scope fence matters in a second sense. Bypass in Section~\ref{sec:results} is scored by \emph{effect}, the marker reaching an out-of-scope network sink, measured \emph{through} the free-text surfaces \(G_s\) leaves open (\texttt{searchFiles.query}, \texttt{respondToUser.message}), not by whether the output is in \(L(G_s)\). The empirical \(0\%\) bypass is therefore not an artifact of those surfaces being closed, and the theorem's guarantee is substantive rather than circular (an anti-circularity argument grounded in the staging-funnel data is given in Section~\ref{sec:res-mistral}). The single-subject realization (Tantalus, a single-subject AI-security agent system) makes this a feasibility result for the covered class, not a population-level universality claim.
\end{remark}

In short: against an adversary who fully controls the logits, GCD's positivity is \emph{enforcement-deep}, reaching all the way down to the generative act. The guarantee is conditional on a correct allowlist and a sound tokenizer lift, both of which we guard with independent, defense-in-depth checks. The detailed proof, with these two obligations promoted to discharged lemmas, is the companion artifact \texttt{docs/proof.md}.
