\section{Limitations and Deviations from Pre-Registration}
\label{sec:limits}

We report these limitations up front, as the pre-registration discipline demands. The three completed datasets establish part of the registered program decisively while leaving named pieces open. We distinguish the two below rather than blur them.

\begin{enumerate}[leftmargin=*]

\item \textbf{Three pretraining lineages, with a single-anchor caveat on the third.} The registered external-validity bar (pre-registration \S12 amendment~\#1: \(\geq 3\) distinct pretraining lineages) is now \textbf{closed}: Qwen (the 1.7B workhorse, the 35B anchor, and the abliterated 27B), Mistral (119B, Sections~\ref{sec:res-mistral} and~\ref{sec:res-ladder}), and Gemma (Gemma-4-31B, Section~\ref{sec:res-gemma}). However, the Gemma lineage rests on a single model/checkpoint (\texttt{nvidia/Gemma-4-31B-IT-NVFP4}): it is one-anchor evidence for that lineage rather than the multi-size coverage Qwen carries. The cached Nemotron-3-Super-120B was abandoned after all gates passed (100\% potency) because its NemotronH hybrid-mamba decode (\(\approx\)64--72 tok/s aggregate, \(\approx\)30$\times$ too slow) made the full ladder infeasible within the available compute allocation.

\item \textbf{The RQ1 headline rides on the Mistral anchor, not the 1.7B.} On Mistral the full stack \texttt{a4} leaks \(1.82\%\) [1.43, 2.29] (CI-lower \(>0\)), Control \(56.77\%\) clears the registered \(\geq 50\%\) potency floor, and the \texttt{a1} \(\leq\) Control manipulation check holds. The 1.7B workhorse is sub-floor \emph{by design} (Control bypass \(18.33\%\); the superseded ungated \texttt{a1} orphan at \(\sim\!20.3\% \geq\) Control fails the A1\(<\)Control check, as expected of a weak victim). It is the weak-end apparatus that sharpens the C-vs-D contrast, not the headline-leak victim.

\item \textbf{Steelman \(N = 2{,}000\)/cell, not the ADR's \(\sim\!10{,}000\).} The abliterated-adversary run (Section~\ref{sec:res-steel}) was scaled down because of eager-mode 27B throughput on a single TP=2 instance. \texttt{c\_sealed} at 0/2000 gives a rule-of-three upper bound of \(\sim\!0.15\%\) (versus \(\sim\!0.03\%\) at 10k). The \(97.85\% \rightarrow 0.00\%\) contrast is unambiguous, but the confidence interval is wider than planned. The \(97.85\%\) control exfil rate (CP95 lower bound \(\approx\!97.1\%\)) is tightly bounded even at \(N=2{,}000\); the wider-interval concern applies only to \texttt{c\_sealed}.

\item \textbf{\texttt{c\_sealed} is partly true-by-construction.} The forced-denylist trajectory makes \(0\) exfiltration structural at the sink, so the \(0.00\%\) is not a discovered safety property of the grammar. Its empirical content is exactly two things: (a) the paired open cell at \(97.85\%\) (\(1957/2000\)), where the adversary is real, willing, and able; and (b) the \(96.8\%\) (\(1937/2000\)) provable forced sensitive-artifact read, the objection-killer against ``it never engaged the malicious content.'' It is reported as a rebuttal exhibit, not main statistics, consistent with ADR~0004. The same structural-ungenerability logic applies in a weaker sense to blind-C's headline result; the distinction is that blind-C is scored by executed effect and the free-text injection surface remained open (see below). The provable read is \emph{engagement evidence}, not a confidentiality result: GCD permits the read and seals only the exfil \emph{action} (confidentiality through free-text channels is Paper-2 scope, Section~\ref{sec:discussion}). The per-skill artifact reads do not generalize uniformly: ``provably read the \emph{AWS credentials}'' (\texttt{AKIAT4NTALUS}\ldots) holds for \(542\); SSH private key \(259\); cross-user chat \(517\); ticket business-data \(619\); the \(63\)-trial remainder are unmapped/multi-message entries that still produced \(0\) exfiltration.

The main-statistics blind-C cells face the same structural-ungenerability fact: an out-of-scope sink URL is also ungenerable by construction, so a naive reading of the headline \(0\)-bypass result is likewise partly definitional. We address this exactly as the pre-registration designed for it (Section~\ref{sec:results}). First, bypass is scored \emph{by effect}: the \texttt{tantalus-pipeline::wins} detector fires only on an executed out-of-scope \texttt{fetchUrl} reaching an attacker sink, not on any grammar-legal token sequence, so the headline figure is ``\(0/N\) executed calls reaching such a sink,'' not ``\(0\) grammar-legal tokens.'' Second, under blind-C the free-text surfaces (\texttt{searchFiles.query}, \texttt{respondToUser.message}) remain open, so an injection route through an open free-text parameter existed and was scored: the attacker could supply arbitrary content in \texttt{searchFiles.query} and \texttt{respondToUser.message}, yet was never observed to produce a bypass.

\item \textbf{The H3 token-cost conjunct fails on the Mistral anchor (\S8).} The registered H3 decision rule is conjunctive: C emission \(=0\) while D emission \(>0\), \emph{and} C's attack-trial output-token cost strictly below D's. The emission and validity-dividend conjuncts carry decisively on both victims: C \(0\) emission versus D \(104\text{k--}120\text{k}\) (1.7B) and \(2{,}317\text{--}3{,}878\) (Mistral). However, the token-cost conjunct \textbf{fails} on the capable Mistral anchor: C at \(208\) output tokens is \emph{not} strictly below \texttt{d\_r1}'s \(203\). So the \emph{formal} conjunctive rule is met \textbf{only} on the 1.7B (C \(119\) \(<\) D \(261\text{--}270\); the ungated \texttt{control} cell at \(289\) tok is not a D variant). The victim-independent result is the qualitative gate-inertness: C makes the corrective gate inert, with \(0\) emission, \(0\) gate-rejection, and \(0\) stranding. Corrective and detective-posture systems, such as Agent-C \cite{agentc2025} and PackMonitor \cite{packmonitor2026}, enforce the same output policy post-generation and face the same produce-then-catch cost. The same produce-then-catch pattern characterises the post-generation policy-enforcement class more broadly (CaMeL \cite{camel2025}, Progent \cite{progent2025}, AgentArmor \cite{agentarmor2025}): the H3 token-conjunct limitation applies to that entire tier, not to any specific implementation, and the strict token ordering is victim-dependent.

The validity dividend (the \(\sim\!41\)\,pp gain in \texttt{valid\_output} on Mistral, Section~\ref{sec:res-mistral}) is not affected by this conjunct failure and is reported separately (Section~\ref{sec:discussion}). It is victim-\emph{dependent} in magnitude: on a high-native-validity model such as Gemma it is negligible (Section~\ref{sec:res-gemma}).

\item \textbf{``0 invalid outputs'' is not literally met on vLLM, and \texttt{c\_guided} is worse than the registered floor.} Per-arm GCD \texttt{valid\_output} on the 1.7B blast: \texttt{c\_d\_r3} \(98.9\%\), \texttt{c\_closed\_d\_r3} \(99.9\%\), \texttt{c\_guided\_d\_r3} \(97.8\%\). The residual is free-string truncation: under the Kleene-\(\ast\) free-string rule on vLLM (the \texttt{\{0,N\}} bound is an \texttt{xgrammar} compile pathology there), a weak model can ramble the open \texttt{respondToUser} message to \texttt{max\_tokens} and truncate. Pre-registration \S12 (angle-shift item~6) registered regular-C \(99.35\%\) and C+ \(99.87\%\). We single out \texttt{c\_guided}'s \(\approx\!2.2\%\) invalid because it is the \textbf{largest} GCD invalidity in this dataset, a fresh disclosed deviation \emph{above} the registered invalid-rate (its terminal reply is the still-open free respond-only string; the fix is a closed canned terminal, as C+ already applies to \texttt{respondToUser}). In fairness, \texttt{c\_d\_r3} at \(98.9\%\) is \emph{also} fractionally below the registered regular-C \(99.35\%\) figure (same Kleene-\(\ast\) free-string cause, same fail-safe behavior), and \texttt{c\_closed\_d\_r3} at \(99.9\%\) is at the registered C+ figure. All fail safe: a truncated string yields no executable tool call (truncation collapses to ``no action,'' never a wrong action; every such trial is \(0\) win / \(0\) emission), and the rates are still far below the \(\sim\!8\text{--}10\%\) parse-fail of non-grammar arms.

\item \textbf{Blind-C utility (RQ4) is victim-dependent, and the powered non-inferiority test fails even when run.} The weak 1.7B shows a \(\approx\!11\)-pp deficit below Control (blind-C legit \(47.0\%\) vs Control \(58.3\%\)); the capable Mistral anchor shows blind-C legit \(78.85\%\) [72.66, 84.19] \emph{above} Control \(73.56\%\). However, that anchor cell (\(n=208\), \texttt{--legit-rounds 1}, CI half-width \(\sim\!6\) pp) is under-powered. The pre-registered powered \(2\)-pp non-inferiority TOST clustered by \texttt{task\_id} (\S8/\S12, \texttt{--legit-rounds 7} \(\approx 1{,}456\) trials/condition) \emph{was} run, on the Mistral ladder (\texttt{mistral119\_ladder.db}, Section~\ref{sec:res-ladder}): observed C\(-\)Control \(= -1.85\) pp (inside the margin) but cluster-bootstrap (\(B=10{,}000\)) 95\% CI \([-7.07, +3.43]\), half-width \(\approx\!5.3\) pp \(\gg\) the \(2\)-pp margin. Non-inferiority is therefore \textbf{not established even when powered}, a confirmed negative rather than an unmeasured gap. What \emph{is} established is twofold. The blind-C deflection deficit does not reproduce on the capable Mistral anchor (Section~\ref{sec:res-mistral}), suggesting a victim-capability confound, though the causal partition has not been measured. And \texttt{c\_guided}/\texttt{c\_l3\_closed} (RQ6, H6) is the universal resolution: bypass \(0\), deflection \(0\), legit-success \(100.0\%\) (\texttt{c\_guided\_d\_r3}, the 1.7B gated cell, \(624/624\)), \(100.00\%\) [99.75, 100] (\texttt{c\_l3\_closed}, the capable Mistral ladder), and \(100.00\%\) [99.41, 100] (\texttt{c\_l3\_closed}, the Gemma anchor; the Gemma run carries no \texttt{c\_guided} cell). The registered H4 precondition required a victim with \(\geq 90\%\) Control legitimate-task success. The Mistral anchor's Control legitimate success was \(73.56\%\) [67.01, 79.42], below this threshold. The powered TOST was executed and reported anyway: the confirmed-negative verdict stands, but on a victim that did not meet the trigger condition. The H6 resolution (\texttt{c\_l3\_closed} attains \(100\%\) legitimate success) requires per-task pre-enumeration of the authorized action sequence, a grammar-authoring overhead the behavioral arms do not carry. The practical deployment trade-off between blind-C (no per-task authoring, non-inferiority unestablished) and L3-closed (guaranteed utility, per-task specification cost) is not resolved by the current experiments and represents an open engineering question.

\item \textbf{Engine/version variance and reproducibility gaps: the primary validity concern (\S10).} The five datasets span different vLLM builds: the 1.7B blast on vLLM 0.23.0, the steelman on vLLM 0.17.1 (the only Blackwell build that loaded the abliterated NVFP4 hybrid coherently), and the Mistral anchor, the Mistral ladder, and the Gemma anchor on from-source / nightly v1 builds with the \texttt{guidance}/\texttt{llguidance} structured-output backend \cite{willard2023,dong2024xgrammar} in place of \texttt{xgrammar} (ADR~0005; deviation rows below). All five produce C-emission \(=0\). The cross-version footguns (AWQ-Marlin hang on \texttt{sm120}, GDN-TP load hang on 0.23.0, the LARK/\texttt{xgrammar} engine crash, and a corrupt-download episode) are apparatus failures that caused the deviation recorded in \S\ref{sec:registered-deviations} (the guidance-backend switch); they do not affect the C-emission \(=0\) result but are recorded as the reason for three of five datasets using a non-registered grammar backend. The reproducibility gap itself is the named primary validity concern of \S10: \textbf{three of the five finding DBs logged \texttt{engine\_commit = unknown} at runtime} (the 1.7B blast, the Mistral anchor, and the steelman, whose \texttt{ENGINE\_COMMIT} env was unset on those victim launches), and were \textbf{backfilled out-of-band} (tagged \texttt{-oob}) from verified engine image digests in \texttt{harness/PROVENANCE.md} (blast NVFP4 rows \texttt{vllm-openai-0.23.0}, blast BF16 rows \texttt{vllm-therock-0.22.1rc1} for \(299{,}829\) early ROCm trials, Mistral \texttt{vllm-ms4-fromsrc}, steelman \texttt{vllm-openai-0.17.1}). The two \(2026\)-\(06\)-\(24\) closeout DBs (the Mistral ladder \texttt{mistral119\_ladder.db} and the Gemma anchor \texttt{gemma31\_anchor.db}) \emph{did} stamp \texttt{engine\_commit} correctly at runtime, so the gap is confined to the three earlier DBs. \texttt{model\_id} remains a served alias; HF checkpoint revisions are resolved in \texttt{PROVENANCE.md} (e.g.\ Mistral \texttt{d57a94c7}\ldots, blast-NVFP4 \texttt{2ab3a24a}\ldots). Per-trial MTP-setting and seed logging (required by \S6/\S10) is unverified for the three backfilled DBs. Because \S10 names reproducibility \emph{the} primary validity concern, we treat these as \textbf{blocking for paper-grade use}, not tail-end polish; native runtime stamping is required on any further runs beyond the two closeout DBs.

\item \textbf{H5 (grammar-compile time): measured out-of-band, registered cold target not met.} The registered overhead envelope is a p99 grammar-compile of \(< 25\) ms (\S5). It was measured with an out-of-band FSM-compile probe over the real R2 grammar and both production tokenizers (\texttt{harness/h5\_grammar\_compile\_results.md}) \cite{dong2024xgrammar}: blind-C \emph{cold} compile is p50 \(38.96\) ms / p99 \(58.64\) ms on the Qwen-151k vocab and p50 \(38.52\) ms / p99 \(61.24\) ms (max \(65.59\) ms) on the Mistral-131k vocab, \textbf{above} the 25\,ms cold target. Amortized over a run with a single fixed grammar (cache hit), per-trial compile is \(\approx\!0.003\) ms (p99 \(0.0030\) ms, Mistral, \(N=500\)). The verdict is measured / qualified: the letter of the 25\,ms cold target is not met, but the spirit holds amortized, because the confirmatory runs use one fixed grammar whose FSM is compiled once and reused. (The \texttt{c\_l2\_guided} \(\approx\)18.6\,KB alternation is much costlier, cold p99 \(346.6\) ms Qwen / \(535.5\) ms Mistral, but amortizes the same way.)

\item \textbf{No formal-proof claim is carried by the empirical samples.} This report and the experiments establish the empirical/implementation tier only: that behavioral controls genuinely leak at scale and that the implementation matches the theory. The class-level ``eliminates by construction'' claim rests on the soundness argument (Section~\ref{sec:formal}), which quantifies over all logit distributions and so holds against a fully injection-compromised model; the full proof is a companion artifact in preparation. Proof and evidence stay in separate lanes.

\end{enumerate}

\subsection{Registered deviations (pre-registration \S12)}
\label{sec:registered-deviations}

Dated rows were added to the pre-registration change log once results were in, each recorded with its reason per the cardinal rule below.

\begin{itemize}[leftmargin=*]
\item \textbf{2026-06-22 -- \texttt{guidance}/\texttt{llguidance} backend.} Three of five datasets used \texttt{guidance} \cite{willard2023,dong2024xgrammar} in place of \texttt{xgrammar} due to a LARK grammar incompatibility (ADR~0005). Item~8 above and the \S\ref{sec:res-ladder} apparatus deviation give the full rationale.
\item \textbf{2026-06-22 -- \texttt{c\_guided} \texttt{valid\_output} \(97.8\%\) exceeds the registered invalid-rate floor.} The terminal free respond-only string under Kleene \(\ast\) on vLLM truncates on a weak victim; it fails safe (\(0\) win / \(0\) emission). Full discussion in item~6 above.
\item \textbf{2026-06-24 -- \texttt{c\_l2\_guided} dropped from the Mistral ladder (guidance-backend incompatibility).} Under the \texttt{guidance}/\texttt{llguidance} backend \texttt{c\_l2\_guided} produced only \(30.2\%\) \texttt{valid\_output} (\(\sim\!72\%\) empty \texttt{raw\_json} traces, with \texttt{guidance} logging ``grammar rejected tokens [0]''): the \(\approx\!18.6\)\,KB \(100\)-way canned-alternation grammar is a construct \texttt{xgrammar} handles (\(97.8\%\) valid on the 1.7B blast) but \texttt{llguidance} does not. \texttt{c\_l2\_guided} was dropped from RUN2 (the Mistral ladder) before data collection began; the ladder instead carries \(c\), \texttt{c\_l3\_closed}, and Control. \emph{Reason:} this is an apparatus incompatibility, not a grammar-soundness failure, since the same grammar enforces correctly under \texttt{xgrammar}.
\item \textbf{2026-06-24 -- H5 grammar-compile time measured out-of-band.} Cold one-time FSM compile (\(\approx\!59\text{--}61\) ms p99) exceeds the registered \(25\) ms p99 cold target; amortized per-trial is \(\approx\!0.003\) ms. Full discussion in item~9 above.
\end{itemize}

\noindent\textbf{Cardinal rule.} Once the confirmatory run began, any deviation from the pre-registration is recorded in \S12 with a reason; numbers reported here come from the analysis plan fixed in the pre-registration, not from post-hoc choices. This discipline, evaluating a positive, default-deny, complete-mediation control \cite{saltzer1975,anderson1972} at scale under a locked protocol, is what licenses reading the constitutive-vs-corrective contrast as a finding rather than a fishing expedition.
