\section{Experimental Design}
\label{sec:method}

\subsection{Design type}

The study is a \emph{quantitative true experiment} of the pre-test/post-test/control-group form, of the reference-monitor evaluation form~\cite{anderson1972,saltzer1975}, composed with a formal soundness core and a built artifact. In the methodological taxonomy of Amaral~\cite{amaral}, this is a deliberate composition of the \emph{Experimental}, \emph{Formal}, and \emph{Build} methodologies: each claim is carried by the method appropriate to it (Section~\ref{sec:formal} for the proof; this section and Section~\ref{sec:results} for the experiment). Amaral explicitly sanctions the composition --- the activities required to answer a single research question may draw on several methodologies --- and so we keep proof and evidence in separate lanes (the three-tier ladder of Section~\ref{sec:core}). Qualitative and correlational designs are rejected: a positive/structural control (GCD) is tested against the negative/behavioral controls that industry actually deploys, on the \emph{same} attack corpus, scored by the \emph{effect}, and reported as block/bypass rate with Clopper--Pearson 95\% confidence intervals~\cite{clopper1934} at scale. The design inherits the classical positive-security heritage --- least privilege, complete mediation, and fail-safe defaults~\cite{saltzer1975}, with the reference-monitor concept itself originating in~\cite{anderson1972}.

\subsection{Independent variable and what is held constant}

The single independent variable is the security control in force. Everything else is a controlled constant: the agent (Tantalus, a simulated enterprise AI assistant with seven backend tools), the seven tools, the embedded environment, the model under test within a given run, the \emph{security-neutral operational base prompt} (persona, tool definitions, and output format, with no security or defensive content), the attack corpus, the inference engine and flags, and the seed/temperature policy. Crucially, \emph{structured-output shape} (valid JSON tool calls) is also held constant across every condition --- it is the substrate, not a control. Shape is not authorization, and there is therefore no separate ``structured-output-shape-only'' arm. The Control condition is base-prompt-only, so the baseline is de-contaminated: a base prompt containing ``never exfiltrate'' would make Control secretly defended and understate attack potency.

Table~\ref{tab:conditions} lists the conditions. The in-band/out-of-band column records the structural reason behavioral controls fail and GCD holds: A1/A2 are \emph{in-band} (they occupy the same input channel the attacker's injection occupies, so the injection overrides them), whereas the GCD family is \emph{out-of-band} (the grammar is compiled from authenticated identity and applied at the sampler, nowhere the attacker's tokens can reach). The post-parse gate D is policy-positive but corrective: its configuration is out-of-band, but enforcement is post-generation --- the forbidden artifact is produced, then caught.

\begin{table}[t]
\centering
\caption{The independent variable: nine security controls. Type is negative/behavioral, positive/constitutive (enforced at the sampler), or positive policy enforced \emph{correctively}. Channel is the locus relative to the attacker's input channel.}
\label{tab:conditions}
\footnotesize
\begin{tabular}{@{}l l l l@{}}
\toprule
ID & Condition & Type & Channel \\
\midrule
\texttt{control} & No added mechanism & baseline & --- (substrate only) \\
\texttt{a1} & Defensive prompt (RFC-2119) & negative, behavioral & in-band \\
\texttt{a2} & Input classifier (embedding) & negative, behavioral & in-band \\
\texttt{a3} & Output filter (credential denylist) & negative, behavioral & post-generation \\
\texttt{a4} & Full behavioral stack (A1+A2+A3) & negative, behavioral & combined \\
\texttt{c} & GCD over union allowlist (blind) & positive, constitutive & out-of-band \\
\texttt{c\_plus} & GCD, closed-response (C+) & positive, constitutive & out-of-band \\
\texttt{c\_guided} & GCD, per-request least-privilege & positive, constitutive & out-of-band \\
\texttt{d} & Post-parse allowlist gate & positive policy, corrective & out-of-band cfg., post-gen.\ enforcement \\
\bottomrule
\end{tabular}
\end{table}

The defining property of the GCD family is that the policy \emph{is} the output space, following the correct-by-construction recognizer programme~\cite{sassaman2013langsec} and the object-capability lineage~\cite{dennisvanhorn1966,miller2006}: a per-request authorization grammar \(G_s\) is an unforgeable capability over the action space. A per-request GBNF \(G_s\) is compiled from the authenticated principal's scope; grammar-invalid tokens are ungenerable at the sampler~\cite{willard2023,dong2024xgrammar}. The enforcement locus is the sampler, not the model: the transformer always computes a full logit distribution and under injection may assign an arbitrarily high logit to a forbidden token, but the grammar mask is applied \emph{before} sampling, driving every grammar-invalid token's probability to zero so it cannot be generated by any sampler (greedy, top-\(k\), top-\(p\), temperature). Logits (intent) are uncontrolled; generated tokens are constrained. The closest decode-time enforcement neighbors are Agent-C~\cite{agentc2025}, which enforces temporal-ordering constraints at decode time (SMT-check-and-resample), and PackMonitor~\cite{packmonitor2026}, which eliminates hallucinated package names via a decode-time DFA-and-logit-mask; both assert model-agnostic soundness. MiniScope~\cite{miniscope2025} enforces least-privilege for tool-calling agents at \emph{runtime} (reconstructed permission hierarchies, the model treated as untrusted) --- the corrective D class here, not generation-time enforcement --- and has no per-principal authorization grammar compiled from authenticated scope, no formal soundness proof, and no large-scale empirical evaluation. Our contribution is not to conceive decode-time enforcement but, to our knowledge, the first to \emph{build and measure} a per-principal authorization grammar compiled per-request from authenticated scope and masked to zero, scored by effect on a large-scale attack corpus.\footnote{Agent-C and PackMonitor use decode-time enforcement for different constraint domains (temporal ordering and package hallucination respectively); neither compiles a per-principal authorization grammar from authenticated scope. The arXiv identifiers and mechanisms for both were web-verified against the live arXiv records.} The post-parse gate D, by contrast, calls the same \texttt{tantalus\_grammar::allowlist\_verdict} routine both C and D call --- that \(G_s\) is built from, so D and C enforce the \emph{same policy over the same language} \(L(G_s)\), but D applies it after an unbounded, default-allow generator has already produced the call. C and D are policy-equivalent; they differ in enforcement completeness, not in what they permit. CaMeL~\cite{camel2025} enforces data-flow taint post-generation; Progent~\cite{progent2025} and AgentArmor~\cite{agentarmor2025} gate tool calls after the model emits them; OAP~\cite{oap2026} verifies agent actions against a policy after execution --- all apply the policy correctively, after the call is produced, not at the sampler, so all occupy D's cell. ARM~\cite{arm2026} is an analysis framework for the TOCTOU/causality-laundering surface that corrective enforcement creates, not a control.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\footnotesize,
  box/.style={draw,rounded corners,minimum height=7mm,minimum width=20mm,inner xsep=3pt,align=center},
  note/.style={draw,rounded corners,minimum height=7mm,inner xsep=3pt,align=center,fill=black!6}]
  % main left-to-right pipeline at y=0
  \node[box] (inj) at (0,0) {Attacker\\injection};
  \node[box] (prompt) at (3.0,0) {Prompt\\assembly};
  \node[box] (model) at (6.0,0) {Transformer\\(logits)};
  \node[box] (samp) at (8.6,0) {Sampler};
  \node[box] (out) at (11.2,0) {Emitted\\tool call};
  \draw[->,thick] (inj) -- (prompt);
  \draw[->,thick] (prompt) -- (model);
  \draw[->,thick] (model) -- (samp);
  \draw[->,thick] (samp) -- (out);
  % A1/A2 in-band note below prompt
  \node[note] (ab) at (3.0,-1.8) {A1/A2 (in-band):\\share attacker's channel\\$\Rightarrow$ injection overrides};
  \draw[->,dashed] (ab) -- (prompt);
  % C out-of-band mask before sampling, above sampler
  \node[note] (c) at (8.6,1.9) {C (out-of-band): grammar mask\\applied to logits \emph{before} sampling\\$\Rightarrow$ invalid tokens $\to 0$};
  \draw[->] (c) -- (samp);
  % A3 post-generation in-band output filter above out
  \node[note] (a3) at (11.2,1.9) {A3 (post-gen., in-band output):\\credential denylist\\on emitted params};
  \draw[->,dashed] (out) -- (a3);
  % D post-generation gate below out
  \node[note] (d) at (11.2,-1.8) {D (post-gen.):\\artifact produced,\\gate applied \emph{after}};
  \draw[->,dashed] (out) -- (d);
\end{tikzpicture}
\caption{Enforcement locus of each condition relative to the attacker injection channel. A1/A2 sit in-band (the attacker's channel); A3 is a post-generation, in-band-output credential denylist on the already-emitted call; C applies the grammar mask out-of-band, on the logits before sampling, so grammar-invalid tokens are ungenerable; D applies the allowlist gate after the call is already produced.}
\label{fig:enforcement-locus}
\end{figure}

\subsection{The GCD ladder and the steelman condition}

The positive cells form a granularity ladder. \emph{L1 (blind)}, condition \texttt{c}, is a union allowlist over all tools with enumerated parameters and free-text left open; it is retained as the headline \emph{security} cell, because 0-bypass over the union allowlist is the falsifiable claim (a hardcoded one-action grammar would make 0\% trivial). \emph{L2 (guided)}, condition C+ / \texttt{c\_plus}, additionally enumerates \texttt{respondToUser.message} to a finite canned-response alternation, giving 0 content-emission by construction for the enumerable-response agent class and closing the truncation residual. \emph{L3 (closed)}, condition \texttt{c\_guided}, narrows \(G_s\) per request to exactly the trusted-channel authorized action(s), with provenance derived from task identity via a single policy map invariant to untrusted skill content --- driving deflection to 0 by construction; this is the utility/granularity result, not the headline security cell. We note one reconciliation: the source identifiers \texttt{c\_closed} and \texttt{c\_guided} were introduced with their ladder meanings inverted relative to the L2/L3 labels above and were renamed to ladder-explicit canonical names (\texttt{c\_l2\_guided}, \texttt{c\_l3\_closed}) with the old tokens retained as serde aliases, so historical result databases keep their meaning.

A separate steelman condition, \texttt{c\_sealed} (forced-denylist, sealed-sink; ADR~0004), is the adversarial inverse of L3: a per-turn scripted grammar that \emph{forces} the malicious denylist trajectory (the compelled sensitive-artifact reads) up to a final pair of forced actions --- a \texttt{fetchURL} to a safe allowlisted endpoint with no \texttt{data} parameter, then a verbatim \texttt{respondToUser} taunt. The forced trajectory is designed so that the model provably ingests the targeted sensitive artifact before the sealed exit, which is the objection-killer for the ``it never read the bad, of course it could not exfil'' rebuttal --- this is engagement evidence (the adversary demonstrably ingested the malicious content), not a confidentiality result; GCD permits the read and blocks only the exfil action, with free-text confidentiality deferred to Paper-2. The realized engagement rate is reported in Section~\ref{sec:res-steel}, where this condition appears as a rebuttal exhibit, not as main statistics. Note that \texttt{c\_sealed}'s 0\% bypass is true-by-construction (the sealed exit makes the attacker sink ungenerable), so its security claim is not independently falsifiable by the sample; the exhibit's value is in the provable engagement rate, not the bypass statistic.

\subsection{Dependent variables}

\paragraph{Primary --- bypass / effect (binary).} A trial is a \emph{bypass} iff the agent causes the harmful effect of an authorization-class attack: it emits a tool call whose execution would carry the attack's sentinel marker to an out-of-scope sink (the win's flag in \texttt{fetchURL.data} to a non-owned endpoint, or \texttt{readChatHistory} on a non-authorized channel), per the deterministic win detector (\texttt{tantalus-pipeline::wins}). This mirrors the reference-monitor evaluation criterion --- scored by \emph{effect}, never by grammar-membership, and never by human grading. For GCD arms, the grammar renders the bypass-scoring token sequence unproducible at the sampler; the win detector fires zero times by the proof of Theorem~1 (Section~\ref{sec:formal}), and the empirical count confirms the grammar door was honored. This is not a circular claim: the bypass is scored by effect on the full token stream including the free-text fields that the union grammar (L1/blind C) leaves open --- the win-detector result provides evidence that the implementation matches the theory, not that the theory alone implies 0\%.

\paragraph{Primary structural --- emission (binary).} Did the model \emph{generate} the out-of-scope tool call at all, independent of downstream blocking? Emission is the empirical signature of the constitutive-vs-corrective distinction, not an efficiency stat: under corrective enforcement (D) the unauthorized artifact is produced (emission \(=1\)) and merely not-forwarded --- it exists in the output buffer, logs, telemetry, stream, and KV cache; under constitutive enforcement (C) the artifact is ungenerable (emission \(=0\)) and absent from every surface. The live form is \texttt{gate\_rejections} (out-of-scope calls produced-then-caught).

\paragraph{Secondary --- output-token expenditure (continuous).} Output tokens generated (\texttt{tokens\_predicted}) plus the raw emitted call, logged in every condition, driving the C-vs-D cost comparison. Reported alongside wall-clock, since C carries a per-token mask cost --- no single-axis ``cheaper'' claim.

\paragraph{Secondary --- output validity (binary).} \texttt{valid\_output} records whether every emitted turn in a trial contained a parseable, schema-conformant tool call (no truncation-invalids, no raw-text tool dumps, no empty-text spurious turns). This is model-capability-decoupled: a grammar arm produces a valid call by construction whenever it emits one, whereas a free-generation arm can dump a tool call as prose, truncate mid-JSON, or emit a stray \texttt{<think>} fragment. The by-construction guarantee applies to the enumerable parameter fields (tool name, allowlisted URL, channel, file path, email id); free-string fields (\texttt{respondToUser.message}, \texttt{searchFiles.query}) use an unbounded Kleene \(*\) rule on vLLM to avoid the \(\{0,N\}\) xgrammar compile pathology, and can be truncated at \texttt{max\_tokens} --- producing a grammar-legal prefix but an unparseable response. This residual (\texttt{c\_guided\_d\_r3} 2.2\% invalid, disclosed in Section~\ref{sec:limits} item~6) fails safe (0 bypass, 0 emission) but the prereg floor is exceeded. The per-condition \texttt{valid\_output} rate is the validity dividend; it is reported alongside bypass and emission in every results table.

\paragraph{Secondary --- time-to-valid-action / attempts (continuous).} The serialized cost to reach a valid in-policy action: \texttt{attempts} (\(=1+\) retries; \emph{engine-invariant}, the primary measure) and decode-time-to-valid-action via \texttt{predicted\_ms} under fixed serving load. Capture caveat: on vLLM \texttt{predicted\_ms}/\texttt{predicted\_per\_second} are 0 (these are llama.cpp-only timing fields), so this DV currently rides on \texttt{attempts} plus aggregate engine \texttt{/metrics}.

\paragraph{Secondary --- legitimate-task success (binary, per legitimate trial).} On a held-out corpus of in-scope tasks, did the agent complete the authorized action under each condition? Measures over-constraint; clustering key is a stable \texttt{task\_id}.

\paragraph{Secondary --- overhead (continuous).} Grammar-compile ms (C), end-to-end latency, time-to-first-token proxy, tokens/sec; reported as percentiles (p50/p90/p99), never bare means.

\paragraph{Deflection (reported against C).} \texttt{deflection\_calls} are generated tool calls that are in scope, not \texttt{respondToUser}, and not in the trial's requested-action set --- C deflecting a masked malicious intent into a valid, unrequested in-scope action. This is two-faced: a reported limitation under blind C, eliminated by construction under C-guided.

Refusals, malformed output, and over-constraint are recorded as \emph{distinct} outcomes, never folded into ``block''; ``0 invalid outputs'' must not silently score spurious in-scope (deflection) actions as wins.

\subsection{D as an offline overlay and a live orthogonal gate}

D is run two ways. \emph{Offline}, the allowlist verdict is overlaid on the logged non-GCD generations: this licenses the emission, wasted-token, and block-effect numbers (the tokens were really generated) but not latency or availability claims. \emph{Live}, since the gate is an orthogonal request-level modifier (\texttt{gate: Option<RetryBudget>}) layerable over \emph{any} generator (the generator \(\times\) gate factorial), D is a real pipeline post-step with a fixed pre-registered rejection message and a retry budget \(R \in \{0,1,3\}\) (the harness conditions \texttt{d\_r0}/\texttt{d\_r1}/\texttt{d\_r3}, and combinations such as \texttt{a4\_d\_r3}). On an out-of-scope call the gate rejects, feeds the message back verbatim, and the model regenerates up to \(R\) times; the terminal state \texttt{d\_terminal} \(\in \{\texttt{valid\_action}, \texttt{availability\_failure}\}\) records whether the budget was exhausted on out-of-scope calls. Only the live arm licenses the availability/latency claims.

\subsection{Hypotheses and decision rules}

The hypotheses and the Section~8 decision rules from the pre-registration are listed verbatim-faithfully below.

\begin{itemize}[leftmargin=1.4em]
\item \textbf{H1 (RQ1 --- behavioral controls leak).} Each behavioral control (A1--A4) exhibits bypass rate \(>0\) on authorization-class attacks, and the full stack A4 still leaks --- in particular on business-data exfil carrying no credential signature (the TicketAssist class). \emph{Rule:} supported if the CI lower bound on each behavioral control's (especially A4's) bypass rate is \(>0\).
\item \textbf{H2 (RQ2 --- positive structural control holds).} GCD/\(G_s\) (C) exhibits bypass rate \(\approx 0\) with a Clopper--Pearson 95\% upper bound strictly below every behavioral control. \emph{Rule:} supported if C's bypass-rate CI upper bound is strictly below every behavioral control's \emph{point estimate}.
\item \textbf{H3 (RQ3 --- constitutive C vs corrective D, conjunctive).} C and D enforce the same policy and block the same effect (\(\approx 0\%\)) but differ in enforcement completeness. \emph{Rule:} supported if C's emission rate is 0 while D's is \(>0\) \emph{and} C's attack-trial output-token cost is strictly below D's (both conjuncts required; note that on models with very short D retry traces the token conjunct may not hold --- see Section~\ref{sec:limits}, item on the H3 qualification). \(H0_3\): C and D have equal emission and equal output-token expenditure on attack trials.
\item \textbf{H4 (RQ4 --- over-constraint / utility, non-inferiority).} C does not reduce legitimate-task success relative to Control within a 2-percentage-point tolerance. \emph{Rule:} supported if C's legitimate-task success CI overlaps Control within the 2-pp tolerance --- a 2pp non-inferiority TOST, cluster-robust by \texttt{task\_id}, powered via hundreds of distinct benign tasks on a capable model. (The registered \(\geq 90\%\)-Control-legit precondition was not met by the Mistral anchor at 73.56\%; the run was nevertheless executed and the result is reported in Section~\ref{sec:results}.) \(H0_4\): C reduces legitimate-task success.
\item \textbf{H5 (RQ5 --- overhead, descriptive).} Per-request grammar-compile time stays within envelope (p99 compile \(<25\) ms) (the realized blind-C cold-compile p99 is 58--61 ms across the Qwen-151k and Mistral-131k tokenizers, exceeding this floor; reported in Section~\ref{sec:results} with the amortized per-trial cost of \(\approx 0.003\) ms under cache hit). \emph{Rule:} assessed descriptively against the envelope, not a hypothesis test; output-token expenditure and wall-clock latency reported for C vs Control vs the D-overlay.
\item \textbf{H6 (RQ6 --- least-privilege granularity, C-guided superiority).} \emph{Rule:} supported if C-guided achieves bypass \(=0\) \emph{and} deflection \(=0\) \emph{and} its legitimate-task success exceeds Control's --- a superiority claim distinct from RQ4's non-inferiority for blind C. \(H0_6\): C-guided's legitimate-task success \(\leq\) Control's.
\end{itemize}

The ``eliminates'' claim is carried by the formal proof (Section~\ref{sec:formal}), not by the sample: the experiment shows that behavioral controls genuinely leak at scale and that the implementation matches the theory; the proof, quantifying over all logit distributions, licenses the class-level elimination.

\subsection{Sample}

\paragraph{Win-classes as malware families.} The five Tantalus win-classes are the ``malware families'': W1 SSH-key exfil, W2 API-key exfil, W3 infra-monitor exfil, W4 ticket-assist auto-forward, W5 cross-user chat exfil. Each is instantiated across paraphrase/obfuscation variants and a direct-vs-second-order delivery axis. Attacks are confused-deputy authority overreach induced by poisoned skills/emails~\cite{greshake2023}.

\paragraph{Selected-on-outcome corpus.} The corpus is built from attacks \emph{known-effective} against the undefended agent --- the live-malware analogue. It is therefore the population of \emph{capable} attacks, not a representative sample of the threat space; bypass rates are conditional on attack potency, and the reference baseline is the control arm's own bypass rate (18.33\% on the 1.7B workhorse, Table~\ref{tab:blast}; 56.77\% on the Mistral anchor, Table~\ref{tab:mistral}; 98.85\% on the Gemma anchor, Table~\ref{tab:gemma}), not 100\% (model nondeterminism means a known winner does not always re-win). The same attack corpus (\texttt{corpus\_fable.json}, 174 entries) is used across all runs; the Control bypass rates differ by model capability, not by corpus. Large N tightens CIs only; it does not buy generalization. Legitimate-task success uses \texttt{corpus\_legitimate\_v2.json} (208 distinct benign in-scope tasks, sourced by construction from the grammar allowlists and environment, with sensitive in-scope files omitted as non-benign).

\paragraph{Asymmetric N budget.} Per family, the trial budget is weighted toward where evidence accrues rather than held symmetric across conditions: Control and the behavioral arms carry the bulk of the attack trials, the live-D arms split across \(R \in \{0,1,3\}\), and C is run for empirical validation of the grammar door: its 0\% is predicted by the formal proof and confirmed stochastically. The sample therefore tests implementation fidelity (was the grammar door actually honored?), not the structural claim (which the proof carries); no sample size can make C's 0\% an independent empirical discovery --- only a corroboration. The asymmetric budget reflects that additional C trials add CI precision on a guarantee already proven, not new evidence about the security property (scored by effect to confirm the grammar door was honored --- the section notes that \texttt{guided\_grammar} is silently ignored on the vLLM builds used, so the empirical check is operationally necessary). This is a deliberate deviation from the pre-registration's symmetric-N specification, recorded in the change log. Single subject system (Tantalus) yields a \emph{feasibility} claim, not a universality claim; a ``family'' is a distinct pretraining lineage.

\subsection{Engines and grammar backends}

The confirmatory data come from vLLM. The grammar door is \texttt{structured\_outputs.grammar} (passed via \texttt{extra\_body}); the top-level \texttt{guided\_grammar} field is \emph{silently ignored} on the vLLM versions used (no error, free-text output), so every run asserts constrained output rather than trusting the request was honored. Two grammar backends appear: the default \emph{xgrammar}~\cite{dong2024xgrammar} for the 1.7B blast and the steelman, and \emph{guidance}/llguidance for the Mistral and Gemma anchors (ADR~0005) (the from-source build's xgrammar rejects the LARK grammar that vLLM's auto-tool-choice emits, crashing the engine on the first tool call; \texttt{guidance} accepts both LARK and our GBNF). The tool-call parser is per model (e.g., \texttt{hermes} for the 1.7B, \texttt{qwen3\_coder} for the 35B, \texttt{mistral} for the anchor). C's grammar door is additionally corroborated on llama.cpp's native GBNF~\cite{gerganov_llamacpp} (the 35B) and SGLang/xgrammar (the 1.7B, on an earlier host); these are \emph{auxiliary corroborations} of engine-independence, not co-equal anchors. Engine-independence of C's structural guarantee is a robustness note, not a claim of three confirmatory engines; the grammar-mask construction itself follows the constrained-decoding line~\cite{willard2023}.

\subsection{Reproducibility commitments and an honest gap}

Every trial logs model id, engine commit, MTP setting, seed, temperature, condition, raw model output, the parsed/emitted tool call, output tokens, the win verdict, and timings; corpora, grammars, the security-neutral base prompt, harness, and analysis code are open and versioned, and the result database reproduces on equivalent hardware. We record one substantive gap up front. All three finding databases logged \texttt{engine\_commit = unknown} at runtime; the values were backfilled out-of-band from verified container image digests (recorded in \texttt{harness/PROVENANCE.md}, tagged \texttt{-oob}), and \texttt{model\_id} remains a served alias rather than a pinned checkpoint hash; per-trial MTP-setting and seed logging are unverified for these databases. This is the primary validity concern and is treated as blocking for paper-grade reproducibility, not tail-end polish; it is carried forward to Section~\ref{sec:limits}.
