% =====================================================================
%  References.
%  The recent-preprint entries below were re-grounded against the live
%  arXiv records (titles, full author lists, venues/years, arXiv IDs)
%  via a web-verification pass (researcher + independent adversarial
%  verifier, each fetching the live arXiv page); each was confirmed to
%  exist. MiniScope (arXiv:2512.11147) was confirmed to exist but found
%  to be a RUNTIME (post-hoc) least-privilege layer, so it is cited only
%  as a Condition-D-class peer (discussion/related), not as
%  generation-time prior art -- see the note on that entry.
% =====================================================================
\begin{thebibliography}{99}

% ---- foundational security-principle lineage ----
\bibitem{saltzer1975}
J.~H.~Saltzer and M.~D.~Schroeder.
\newblock The protection of information in computer systems.
\newblock \emph{Proceedings of the IEEE}, 63(9):1278--1308, 1975.

\bibitem{anderson1972}
J.~P.~Anderson.
\newblock Computer security technology planning study.
\newblock Technical Report ESD-TR-73-51, U.S.\ Air Force Electronic Systems Division, 1972.
\newblock (Origin of the reference-monitor concept.)

\bibitem{sassaman2013langsec}
L.~Sassaman, M.~L.~Patterson, S.~Bratus, and M.~E.~Locasto.
\newblock Security applications of formal language theory.
\newblock \emph{IEEE Systems Journal}, 7(3):489--500, 2013.
\newblock (The LANGSEC programme: correct-by-construction recognizers.)

\bibitem{dennisvanhorn1966}
J.~B.~Dennis and E.~C.~Van~Horn.
\newblock Programming semantics for multiprogrammed computations.
\newblock \emph{Communications of the ACM}, 9(3):143--155, 1966.

\bibitem{miller2006}
M.~S.~Miller.
\newblock \emph{Robust Composition: Towards a Unified Approach to Access Control and Concurrency Control}.
\newblock PhD thesis, Johns Hopkins University, 2006.
\newblock (Object-capability security.)

% ---- the threat and the benchmark landscape ----
\bibitem{greshake2023}
K.~Greshake, S.~Abdelnabi, S.~Mishra, C.~Endres, T.~Holz, and M.~Fritz.
\newblock Not what you've signed up for: Compromising real-world LLM-integrated applications with indirect prompt injection.
\newblock In \emph{Proc.\ ACM Workshop on Artificial Intelligence and Security (AISec)}, 2023.

\bibitem{zhang2024asb}
H.~Zhang, J.~Huang, K.~Mei, Y.~Yao, Z.~Wang, C.~Zhan, H.~Wang, and Y.~Zhang.
\newblock Agent Security Bench (ASB): Formalizing and benchmarking attacks and defenses in LLM-based agents.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2025.
\newblock arXiv:2410.02644.

% ---- nearest neighbours: decode-time / generation-time enforcement ----
\bibitem{agentc2025}
A.~Kamath, S.~Zhang, C.~Xu, S.~Ugare, G.~Singh, and S.~Misailovic.
\newblock Enforcing temporal constraints for LLM agents.
\newblock arXiv:2512.23738, 2025. \url{https://arxiv.org/abs/2512.23738}.

\bibitem{packmonitor2026}
X.~Liu, Y.~Liu, Y.~Zhang, J.~Li, and S.-M.~Hu.
\newblock PackMonitor: Enabling zero package hallucinations through decoding-time monitoring.
\newblock arXiv:2602.20717, 2026. \url{https://arxiv.org/abs/2602.20717}.

% ---- positive / post-generation cell (Condition D's peers) ----
\bibitem{camel2025}
E.~Debenedetti, I.~Shumailov, T.~Fan, J.~Hayes, N.~Carlini, D.~Fabian, C.~Kern, C.~Shi, A.~Terzis, and F.~Tram\`er.
\newblock Defeating prompt injections by design (CaMeL).
\newblock arXiv:2503.18813, 2025. \url{https://arxiv.org/abs/2503.18813}.

\bibitem{progent2025}
T.~Shi, J.~He, Z.~Wang, H.~Li, L.~Wu, W.~Guo, and D.~Song.
\newblock Progent: Securing AI agents with privilege control.
\newblock arXiv:2504.11703, 2025. \url{https://arxiv.org/abs/2504.11703}.

% Web-verified to exist (arXiv:2512.11147; authors/title/year confirmed
% against the live arXiv record). MiniScope is a RUNTIME / post-hoc
% least-privilege layer (reconstructed permission hierarchies, LLM
% treated as untrusted) -- the corrective Condition-D class -- NOT a
% grammar/sampler/decoding-time constraint. Now cited ONLY in that role
% (discussion/related); the earlier generation-time attribution was
% corrected out of abstract/intro/method/formal (2026-06-24).
\bibitem{miniscope2025}
J.~Zhu, K.~Tseng, G.~Vernik, X.~Huang, S.~G.~Patil, V.~Fang, and R.~A.~Popa.
\newblock MiniScope: A least privilege framework for authorizing tool calling agents.
\newblock arXiv:2512.11147, 2025. \url{https://arxiv.org/abs/2512.11147}.

\bibitem{fides2025}
M.~Costa, B.~K\"opf, A.~Kolluri, A.~Paverd, M.~Russinovich, A.~Salem, S.~Tople, L.~Wutschitz, and S.~Zanella-B\'eguelin.
\newblock Securing AI agents with information-flow control (Fides).
\newblock arXiv:2505.23643, 2025. \url{https://arxiv.org/abs/2505.23643}.

\bibitem{agentarmor2025}
P.~Wang, Y.~Liu, Y.~Lu, Y.~Cai, H.~Chen, Q.~Yang, J.~Zhang, J.~Hong, and Y.~Wu.
\newblock AgentArmor: Enforcing program analysis on agent runtime trace to defend against prompt injection.
\newblock arXiv:2508.01249, 2025. \url{https://arxiv.org/abs/2508.01249}.

\bibitem{oap2026}
U.~Uchibeke.
\newblock Before the tool call: Deterministic pre-action authorization for autonomous AI agents.
\newblock arXiv:2603.20953, 2026. \url{https://arxiv.org/abs/2603.20953}.

\bibitem{arm2026}
M.~H.~Chinaei.
\newblock Causality laundering: Denial-feedback leakage in tool-calling LLM agents.
\newblock arXiv:2604.04035, 2026. \url{https://arxiv.org/abs/2604.04035}.

% ---- grammar-constrained / structured generation machinery ----
\bibitem{willard2023}
B.~T.~Willard and R.~Louf.
\newblock Efficient guided generation for large language models.
\newblock arXiv:2307.09702, 2023.

\bibitem{dong2024xgrammar}
Y.~Dong, C.~H.~Yin, et~al.
\newblock XGrammar: Flexible and efficient structured generation engine for large language models.
\newblock arXiv:2411.15100, 2024.

\bibitem{gerganov_llamacpp}
G.~Gerganov and contributors.
\newblock llama.cpp: LLM inference in C/C++ with GBNF grammar-constrained sampling.
\newblock Software. \url{https://github.com/ggml-org/llama.cpp}.

% ---- adversarial upper bound ----
\bibitem{arditi2024}
A.~Arditi, O.~Obeso, A.~Syed, D.~Paleka, N.~Panickssery, W.~Gurnee, and N.~Nanda.
\newblock Refusal in language models is mediated by a single direction.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2024.

% ---- statistics and methodology ----
\bibitem{clopper1934}
C.~J.~Clopper and E.~S.~Pearson.
\newblock The use of confidence or fiducial limits illustrated in the case of the binomial.
\newblock \emph{Biometrika}, 26(4):404--413, 1934.

\bibitem{amaral}
J.~N.~Amaral et~al.
\newblock About computing science research methodology.
\newblock Methodology primer.

\end{thebibliography}
