\section{Related Work}
\label{sec:related}

We organize the prior art along the two axes that distinguish the controls in this study: the \emph{security model} (negative, enumerate and detect bad behavior, versus positive, enumerate and admit only authorized behavior under default-deny) and the \emph{enforcement stage} (post-generation, after the model has produced an artifact, versus at-generation, at the decoding boundary). The resulting \(2\times2\), summarized in Table~\ref{tab:related-2x2}, locates every control class and exposes the cell this work occupies. The positive-versus-negative contrast we draw, enumerate-and-admit-the-authorized against enumerate-and-detect-the-bad scored by effect at scale, recapitulates the classical host-security shift from signature antivirus to application allowlisting. However, its intellectual roots lie in a longer lineage of positive-security design that we credit next.

\subsection{Foundational lineage}
\label{sec:related-lineage}

The positive-security stance this work operationalizes descends from four classical lines. Saltzer and Schroeder~\cite{saltzer1975} established \emph{complete mediation} and \emph{fail-safe defaults} (default-deny) as design principles: every access is checked against an explicit policy, and the absence of an explicit permission denies. The \emph{reference-monitor} concept~\cite{anderson1972} gave this an architectural form, a tamper-proof, always-invoked, verifiable mediator of every access, which the post-generation positive controls in our matrix (Condition \texttt{d}) approximate at execution time without the always-invoked guarantee, since the artifact exists before the gate fires. The LangSec programme~\cite{sassaman2013langsec} contributed the \emph{correct-by-construction recognizer}: when the set of admissible inputs is a formal language, a recognizer for that language enforces correctness by making malformed inputs unrepresentable rather than detectable. This is the direct intellectual precursor to grammar-as-policy, where the language \emph{is} the policy. Finally, the object-capability model~\cite{dennisvanhorn1966,miller2006} frames authority as an unforgeable token over a designated action space, held only by principals entitled to it. Most directly, Galloway~\cite{galloway2020} gave this lineage an empirical control-evaluation form in host security: a controlled comparison of application allowlisting (a positive, default-deny control) against signature-based malicious-code detection (a negative control). The present study ports that evaluation design into the AI-security domain, substituting the LLM agent's tool-call action space for the host's executable space and the behavioral guardrails for signature detection. Grammar-constrained decoding (GCD) over a per-request authorization grammar \(G_s\) sits at the intersection: \(G_s\) is the reference monitor relocated to the sampler, compiled per request from authenticated caller identity (an unforgeable capability over the agent's action space) and realized as a correct-by-construction recognizer at the decoding boundary. Among the layers this study adds is a formal soundness core (Section~\ref{sec:formal}): constrained generation can emit only strings in the grammar's language, and the soundness statement \(\textit{emitted} \in L(G_s) \subseteq \mathrm{Authorized}(s)\) is quantified over all logit distributions, so it holds against a fully injection-compromised model. That is what licenses a class-level elimination-by-construction claim no finite sample could: the empirical evidence shows that behavioral controls genuinely leak at scale and that the implementation matches the theory.

\subsection{The threat: indirect prompt injection and the benchmark landscape}
\label{sec:related-threat}

The attacks we study are predominantly \emph{indirect} prompt injections delivered through a trusted data channel rather than the user message. Greshake et al.~\cite{greshake2023} characterized this class: malicious instructions ride inside content the agent retrieves and trusts (documents, emails, web pages), so the injection occupies the same input channel as any legitimate instruction. This is precisely why the in-band behavioral controls in our matrix, the defensive system prompt (\texttt{a1}) and the embedding input classifier on the user message (\texttt{a2}), can be overridden: they live in the channel the attacker shares. The agent-security benchmark literature has documented the empirical leakiness of behavioral defenses at scale. The Agent Security Bench~\cite{zhang2024asb} evaluated a broad matrix of attack and defense methods across many models over 90{,}000 test cases, reported high peak attack-success rates, and concluded that prevention-based defenses of the detect-bad class are inadequate. Our confirmatory corpus subjects a single subject system, Tantalus (a locally-deployed AI-security agent), to roughly \(27\times\) that benchmark's test-case count. We stress that this buys tighter confidence intervals, not generalization, because the corpus is selected-on-outcome (the population of capable attacks against the undefended agent) and bypass rates are therefore conditional on attack potency. Bypass is scored by effect: the win detector (\texttt{tantalus-pipeline::wins}) fires only when the exfiltration marker reaches an out-of-scope sink URL in an executed \texttt{fetchUrl} call, not on whether the call is in \(L(G_s)\), so a grammar-constrained call to an allowlisted URL scores zero bypass regardless (Section~\ref{sec:method}).

\subsection{Grammar-constrained and structured generation machinery}
\label{sec:related-machinery}

The mechanism we deploy, constraining the sampler to a formal language, builds directly on the structured-generation line of work. Willard and Louf~\cite{willard2023} cast constrained generation as an efficient finite-state-machine guided decoding problem; xgrammar~\cite{dong2024xgrammar} provides a high-throughput pushdown-automaton token-mask backend for context-free grammars; and \texttt{llama.cpp}~\cite{gerganov_llamacpp} exposes native GBNF grammars at the sampler. Our enforcement rides on this machinery, and the same C-emission-of-zero result reproduces across engines and grammar backends. We treat this cross-engine agreement as an engine-independence robustness corroboration rather than as several co-equal anchors: \texttt{vLLM}-xgrammar (the \(1.7\mathrm{B}\) blast and the steelman) and \texttt{vLLM}-guidance/llguidance (the Mistral anchor) are the confirmatory engines for the reported data, while \texttt{llama.cpp}-GBNF (on the \(35\mathrm{B}\) anchor) and SGLang-xgrammar (on the \(1.7\mathrm{B}\), prior apparatus) serve as auxiliary corroborations. The distinction we draw is one of \emph{purpose}, not of plumbing. The dominant deployed use of this machinery is structured-output \emph{shaping}: forcing output to a JSON schema or fixed shape. Shaping is not authorization: a schema constrains the \emph{form} of a tool call, not \emph{which} URL, channel, or file the call may name. In our design the structured-output shape is held constant across every condition (it is the substrate, not a control). The novelty is per-request \emph{authorization} compiled into the grammar from the authenticated caller identity and applied at the decoder, so the action space \emph{is} the policy and a forbidden parameter value is ungenerable rather than merely mis-shaped.

\subsection{The \texorpdfstring{$2\times2$}{2x2}: where the deployed controls sit, and the empty cell}
\label{sec:related-cells}

\paragraph{Negative, post-generation.}
This cell is the industry stack: defensive prompting, input classifiers, output filters, and LLM-judge / reasoning-monitor guardrails. They are probabilistic detectors of bad behavior applied after (or, for prompt-level defenses, in-band with) generation. Output filters in particular are a credential-\emph{signature} denylist, incomplete by construction: business data that carries no credential signature passes through unflagged (our \texttt{a3} arm, and the survivor class on the capable anchor in Section~\ref{sec:res-mistral}). A recurring real-world hazard in this cell is that several commercial ``agent security'' products market themselves as preventative or allowlist-based while being, mechanically, detectors. This is the motivating gap this study addresses. We name these product classes in prose only; no citation is asserted for them.

\paragraph{Positive, post-generation.}
The academic frontier has moved security into this cell: deterministic, execution-time reference monitors and policy compilers that admit only authorized actions but enforce the policy \emph{after} the model has generated the tool call. The action is produced in full, then a separate gate inspects and rejects it. Several recent systems realize this class. CaMeL~\cite{camel2025} and Progent~\cite{progent2025} enforce authorization post-generation via dual-LLM and programmable privilege-control architectures; MiniScope~\cite{miniscope2025} applies least-privilege permission scoping at tool-call time, enforcing post-generation; FIDES~\cite{fides2025} uses information-flow control (applied to the generated artifact's provenance graph post-production, placing it in this cell rather than at-generation; it extends into Paper-2 scope for free-text channels); AgentArmor~\cite{agentarmor2025} performs program analysis over agent runtime traces; OAP~\cite{oap2026} provides deterministic \emph{pre-execution} authorization (pre-execution, not pre-generation, so the artifact still exists before the gate); and ARM~\cite{arm2026} identifies a risk that reference monitors in this class can leak information through the denial feedback itself, a surface that emission~\(=0\) structurally forecloses. Our Condition \texttt{d} is a faithful instance of exactly this class: it applies the \emph{same} allowlist that \(G_s\) compiles (C and D call the same \texttt{tantalus\_grammar::allowlist\_verdict} routine, so \(D \equiv C\) by construction), but correctively. This is the right comparator, and it is effective at blocking the effect; we do not claim C is more secure than a perfectly wired allowlist on the blocked effect. The two controls enforce the same positive policy over the same language \(L(G_s)\) and differ in enforcement completeness, not in what they permit. The corrective allowlist re-introduces, one layer down, the failure modes positive security exists to escape: the dangerous artifact is produced (a cost and an output-channel / forensics surface, the emission signature), a coverage obligation across every output path and consumer, and a time-of-check/time-of-use window~\cite{saltzer1975}. GCD has none of these because there is nothing to catch. Accordingly we deploy a post-parse check of this class as a defense-in-depth backstop \emph{complementing} C, not as a control C dominates.

\paragraph{Positive, at-generation (this work).}
The remaining cell, positive security enforced at the decoding boundary, was the least occupied prior to this work. The nearest decode-time neighbours are Agent-C~\cite{agentc2025}, which enforces temporal constraints at the sampling boundary with a model-agnostic soundness claim, and PackMonitor~\cite{packmonitor2026}, which eliminates package-hallucination via a decode-time automaton monitor that is the mechanism twin of ours (a DFA driving a logit mask). Both demonstrate the viability of decode-time enforcement but differ in scope: Agent-C constrains action \emph{ordering} (and checks temporal formulae by SMT-and-resample rather than masking the forbidden token to zero), and PackMonitor constrains package-name \emph{enumeration} for hallucination-correctness, whereas this work compiles per-request \emph{authorization} policy from authenticated caller identity (the action space \emph{is} the policy) and masks every out-of-scope token to zero. Agent-C's resample-until-valid approach leaves each individual draw able to produce a constraint-violating token; GCD's mask-to-zero makes every forbidden token ungenerable at any draw, which is where the no-artifact guarantee lives. We therefore claim to be the first to \emph{build and measure} generation-time authorization at scale across multiple model families with formal soundness, not the first to conceive decode-time enforcement. GCD at the sampler moves enforcement one stage earlier than the execution-time monitor: the grammar mask is applied before sampling, driving every grammar-invalid token's probability to zero, so for enumerable-sink parameters the out-of-scope action is ungenerable by any sampler and never exists as an artifact on any surface (output buffer, logs, telemetry, stream, KV cache). Free-text fields (\texttt{searchFiles.query}, \texttt{respondToUser.message}) remain open-vocabulary; free-text confidentiality leakage is out of scope for the headline security claim (Section~\ref{sec:limits}). Enforcement is at the sampler, not the model: the transformer still computes a full logit distribution and under injection may assign an arbitrarily high logit to a forbidden token, but that intent never becomes a generated token. Because the grammar is compiled from authenticated identity and applied at the decoder, it is out-of-band, nowhere the attacker's injected tokens can reach, which is the precise reason the in-band behavioral controls fail and this one holds. This cell is where enforcement is provable and leaves no emission. With xgrammar (which removes the per-token CPU-mask cliff) C is at-par with free decoding in the agentic loop, and the per-token masking cost is reported honestly against the avoided regeneration. A structural corollary is a model-agnostic \emph{validity} guarantee: every emitted token is in \(L(G_s)\) by construction, so valid-output rates are decoupled from model capability. A \(1.7\mathrm{B}\) victim under GCD reaches \(98.9\%\) valid outputs (Section~\ref{sec:res-blast}, \texttt{c\_d\_r3}) versus \(90\%\) under control (\(89.4\)--\(93.9\%\) across behavioral conditions), independent of the model's fluency. However, the validity dividend is victim-dependent: it is largest where native free generation is noisy (\(\approx\)59\% on Mistral, \(89\)--\(94\%\) on the weak \(1.7\mathrm{B}\)) and negligible where the model already produces well-formed tool calls natively (Gemma, Section~\ref{sec:res-gemma}). In both regimes GCD validity is model-agnostic by construction. This is a reliability contribution separate from the bypass result. The enforcement point is also distinctive: the mask is applied at every emitted token rather than after a finished call, so it acts on the free-text and streaming surfaces an execution-time monitor inspects only after the fact, a statement about \emph{where} enforcement happens, scoped to action integrity over enumerable-sink tools.

\begin{table}[t]
\centering
\small
\caption{The design space as a \(2\times2\): security model \(\times\) enforcement stage. Each cell names the deployed control class and its nearest published neighbours; Condition labels in \texttt{typewriter}. The positive/at-generation cell is the contribution of this work.}
\label{tab:related-2x2}
\begin{tabular}{@{}p{0.14\textwidth} p{0.38\textwidth} p{0.38\textwidth}@{}}
\toprule
 & \textbf{Negative} (detect-bad) & \textbf{Positive} (allowlist, default-deny) \\
\midrule
\textbf{Post-generation} & Behavioral / LLM-judge guardrails: defensive prompt (\texttt{a1}), input classifier (\texttt{a2}), credential output filter (\texttt{a3}), full stack (\texttt{a4})~\cite{greshake2023,zhang2024asb}; commercial ``agent security'' marketed as preventative but mechanically a probabilistic detector. & Deterministic reference monitors / policy compilers / execution-time allowlists: CaMeL~\cite{camel2025}, Progent~\cite{progent2025}, MiniScope~\cite{miniscope2025}, FIDES~\cite{fides2025}, AgentArmor~\cite{agentarmor2025}, OAP~\cite{oap2026}, ARM~\cite{arm2026} (Condition \texttt{d} class). Same policy as \(C\); the artifact is produced then caught, so emission, coverage, and TOCTOU surfaces remain. \\
\addlinespace
\textbf{At generation} & ---\textsuperscript{$\dagger$} & \textbf{GCD at the sampler (\texttt{c}/\texttt{c\_guided}), this work.} Nearest neighbours: Agent-C~\cite{agentc2025} (temporal ordering, SMT-resample) and PackMonitor~\cite{packmonitor2026} (package-name DFA), distinguished by per-request authorization scope and mask-to-zero. Provable (Section~\ref{sec:formal}), no emission, at-par with free decoding in the agentic loop (vLLM/xgrammar); the mask applies per-token, not after a finished call (action-integrity scope: enumerable-sink parameters). \\
\bottomrule
\end{tabular}

\vspace{2pt}
{\footnotesize \textsuperscript{$\dagger$}No published system is known to enforce a negative/detect-bad policy at the decoding boundary; the cell is structurally empty because detect-bad and at-generation are jointly incoherent (detection requires comparing against an artifact, which does not yet exist at the decoding boundary).}
\end{table}

\subsection{Abliteration and the adversarial upper bound}
\label{sec:related-abliteration}

Behavioral safety can be surgically removed from a model's weights. Arditi et al.~\cite{arditi2024} showed that refusal behavior in open-weight models is mediated by a single linear direction in the residual stream that can be ablated, yielding a model that cannot refuse. This is the natural steelman for our thesis: abliteration defeats every weight- and prompt-resident behavioral defense (the negative/post-generation cell) by construction, because there is no refusal left to elicit. It cannot, however, touch a constraint that lives in the sampler. Section~\ref{sec:res-steel} reports the resulting contrast on a fully abliterated model under a top-authority malicious prompt: 97.85\% exfiltration with the sink open versus 0.00\% under a sampler-sealed grammar, with the grammar as the sole independent variable. We present this as a rebuttal exhibit rather than a primary statistic, and we are explicit (Section~\ref{sec:limits}) that the sealed cell is true-by-construction. Its empirical content is the paired open cell (the adversary is real, 97.85\% exfiltration) and the \(1{,}937/2{,}000\) (\(96.8\%\)) provably-forced sensitive reads across all mapped skills (the engagement-evidence objection-killer), not a claim that the grammar discovered safety.
