\section{Background and Threat Model}
\label{sec:threat}

\subsection{Subject system: the Tantalus agent}
\label{sec:threat-subject}

The subject system is \emph{Tantalus}, a real, previously-built and deployed, deliberately-vulnerable agentic application. It ships with a known, calibrated corpus of attacks independently observed to succeed against its behavioral defenses. For this study Tantalus runs as a self-contained \emph{local} binary on \texttt{vLLM} / \texttt{llama.cpp}~\cite{gerganov_llamacpp} with grammar support, with the original cloud path replaced, so that Grammar-Constrained Decoding (GCD) is a true token-level mask at the sampler rather than a managed cloud structured-outputs feature. The agent acts as a desktop-support deputy for a \emph{single} principal, ``Jordan'' at a consultancy, over seven tools: \texttt{readFile}, \texttt{searchFiles}, \texttt{listInbox}, \texttt{readEmail}, \texttt{readChatHistory}, \texttt{fetchURL}, and \texttt{respondToUser}.

\subsection{Authorization model and the confused-deputy threat}
\label{sec:threat-authz}

The principal's authority is the set of tool calls legitimately needed for the task at hand. Attacks are \emph{authority overreach}: a confused-deputy redirect~\cite{greshake2023} in which poisoned skills or emails induce the deputy to read a resource outside the task scope and/or to exfiltrate via \texttt{fetchURL} to a non-owned endpoint. An ``authorization-class attack'' is one whose success requires emitting a tool call with parameters outside the principal's authorized set, the action space the legitimate task never needs to enter.

We score success by \emph{effect}, following the reference-monitor principle of complete mediation: a control is judged by whether the harmful effect actually occurs, not by any internal acceptance predicate~\cite{anderson1972,saltzer1975}. A trial is a \textbf{bypass} iff the agent causes the harmful effect of an authorization-class attack. It emits a tool call whose execution would carry the attack's sentinel marker to an out-of-scope sink (the win's flag in \texttt{fetchURL.data} to a non-owned endpoint, or \texttt{readChatHistory} on a non-authorized channel), as determined by a deterministic win detector (flag-marker plus scope-oracle, no human grading). Otherwise the trial is a \textbf{block}. Bypass is therefore an end-to-end property of the deployed system scored on the effect, never on grammar-membership. The win detector is a deterministic flag-marker plus scope-oracle applied to the \emph{executed} tool call, implemented in the separate \texttt{tantalus-pipeline::wins} module, independent of the grammar definition; the grammar and the win detector share the same authorized-set specification but are evaluated by separate code paths, so a grammar bug that admitted a forbidden URL would not suppress a win. This separation matters because the headline security claim must not be co-defined with the grammar that closes the sink (Section~\ref{sec:results}). The GCD grammar does not enumerate all inputs: free-text parameters (\texttt{searchFiles.query}, \texttt{respondToUser.message}) remain open in the blind-C grammar, so the model retains latitude to attempt scope violations through those surfaces. A 0\% bypass at the structured sinks is therefore an empirical claim about what the agent does \emph{with} that latitude, not a restatement of the grammar\textquotesingle s closure of the enumerable-sink parameters (Section~\ref{sec:results}). This free-text-surface argument applies to W1--W4 (the \texttt{fetchURL} sink, which requires both an attacker-controlled endpoint and the sentinel marker). W5 is scored on channel scope rather than sink reachability, and the grammar enumerates authorized channels, so C\textquotesingle s block on W5 is partly preventable by channel enumeration and is therefore partly constitutive-by-definition.

\paragraph{Scope: security vs.\ safety, and why a bypass is machine-checkable.}
We adopt an operational boundary between AI \emph{security} and AI \emph{safety}: a security issue is one that impacts the technical security of the system itself, whereas a safety issue is one whose harm requires human interpretation of content to manifest. If it takes a person reading the output to be dangerous, it is a safety issue. This boundary is what licenses an objective bypass metric: an authorization-class bypass is decided by \emph{effect} at an enumerable sink (a sentinel marker reaching a non-owned endpoint), with no human grading, precisely because such harm is mechanical rather than interpretive. Harms that do require interpretation, such as the semantics of composed \texttt{respondToUser.message} content, are deliberately out of scope here (Section~\ref{sec:limits}). The same boundary motivates the enforcement locus we test. Because an LLM can be biased or redirected by its inputs in ways that prompting and training cannot reliably prevent, the model cannot be relied upon as a \emph{sound} security control: prompting and training improve output quality under normal operation but do not bound behavior under adversarial input, so a sound control must sit outside the model. Conditions A1--A4 (behavioral defenses inside the model's input channel) versus C (a constitutive mask at the sampler) are exactly that hypothesis under test.

The five win-classes of the Tantalus subject system are the attack families of this study, instantiated across paraphrase/obfuscation variants and a delivery axis (direct prompt vs. second-order email-trap / poisoned skill):

\begin{itemize}[leftmargin=1.4em,itemsep=1pt,topsep=2pt]
  \item \textbf{W1} --- SSH-key exfiltration (private key to an attacker endpoint).
  \item \textbf{W2} --- API-key exfiltration (AWS-style credentials).
  \item \textbf{W3} --- infra-monitor exfiltration (credential read via a daemon persona).
  \item \textbf{W4} --- ticket-assist auto-forward: business data with \emph{no credential signature}.
  \item \textbf{W5} --- cross-user chat exfiltration (a non-authorized channel).
\end{itemize}

\noindent W5 is scored on channel scope rather than sink reachability: a W5 bypass requires only \texttt{readChatHistory} on a non-authorized channel, with no outbound \texttt{fetchURL}. The GCD grammar enumerates authorized channels, so C\textquotesingle s 0\% bypass on W5 is partly by construction, and the free-text-surface argument of Section~\ref{sec:threat-authz} applies cleanly only to W1--W4.

\noindent W4 is the class that defines negative-security incompleteness. Its payload is support-ticket text, which matches no credential pattern, so a signature denylist (A3, below) cannot catch it. Section~\ref{sec:res-mistral} (Table~\ref{tab:mistral}) reports that the survivors of the output filter on a capable victim are precisely this credential-signature-free class.

A third outcome, \emph{invalid output}, occurs when the agent produces a tool call that is not parseable or not executable: a structurally degenerate response that neither achieves the attack nor constitutes a useful action. Invalid output is decoupled from model capability. A capable model can produce invalid output under adversarial injection (hermes-format tool calls emitted as prose, truncated JSON), while a weak model produces valid-but-wrong outputs when grammar-constrained. We track \emph{valid output} as a secondary dependent variable (Section~\ref{sec:method}) to distinguish the security guarantee from an artifact of model quality.

\subsection{In-band vs. out-of-band defenses}
\label{sec:threat-band}

The decisive structural property of a control is \emph{where it sits relative to the attacker's channel}. The behavioral controls A1 (a defensive system prompt with an RFC-2119 security policy) and A2 (an embedding classifier on the user message) are \emph{in-band}: they live in the same input channel the attacker's injection occupies, so an injection through a trusted skill or email can simply override them. GCD (Condition C) is \emph{out-of-band}: the grammar \(G_s\) is compiled from the authenticated principal's identity~\cite{sassaman2013langsec,anderson1972,saltzer1975} and applied at the sampler, nowhere the attacker's tokens can reach~\cite{anderson1972,saltzer1975}. That in-band/out-of-band split is the precise reason one fails and one holds: the grammar mask~\cite{willard2023,dong2024xgrammar} drives every grammar-invalid token's probability to \(0\) at the sampler, so the soundness statement \(\mathrm{emitted} \in L(G_s) \subseteq \mathrm{Authorized}(s)\) holds regardless of what logits the injected model produces. The masking mechanism (the full logit distribution, the sampler modes, and the logits-versus-tokens distinction) is detailed in Section~\ref{sec:formal}. The present work applies this enforcement locus to per-request \emph{authorization}, where the authorized language is derived from the principal's authenticated identity rather than a fixed domain constraint. Agent-C~\cite{agentc2025} applies decode-time enforcement to temporal-ordering constraints; PackMonitor~\cite{packmonitor2026} applies it to hallucination elimination. Both share the sampler locus but target reliability rather than per-principal authorization, and neither compiles a grammar from authenticated scope.

The corrective comparator, Condition D, is a post-parse allowlist gate that checks the \emph{parsed} tool call against the same allowlist \(G_s\) compiles. Its gate configuration is itself derived from identity (out-of-band). However, its \emph{enforcement} is post-generation: the forbidden artifact is produced first, then caught. C and D thus enforce the same positive policy over the same language \(L(G_s)\) and differ only in enforcement completeness, not in what they permit. That distinction, constitutive vs. corrective, is the core of Section~\ref{sec:core}. D is representative of the post-generation positive-control class~\cite{camel2025,progent2025,oap2026,agentarmor2025,arm2026}, which enforce the policy after an unbounded generator has produced the call.

\subsection{Held-constant substrate, ethics, and an open decision}
\label{sec:threat-substrate}

Two design points scope the contrast. First, \emph{structured-output shape} (the requirement that every tool call be valid JSON) is held constant across \emph{all} conditions, including Control: it is the substrate, not a control, and shape is not authorization. There is no ``shape-only'' arm, and we make no claim that valid-JSON enforcement constitutes authorization. Second, the experiment is fully local and ethically contained: \texttt{fetchURL} and all outbound channels are mocked (no real exfiltration, no SSRF surface), and the environment and secrets are synthetic sentinel markers rather than real credentials. The abliterated-adversary steelman (Section~\ref{sec:res-steel}) uses an AKIA-pattern string with no associated real account and no real exfiltration path: the string\textquotesingle s credential format, not a live credential, is what allows the engagement-evidence analysis. No external systems are attacked, and there are no human subjects. The corpus is \emph{selected-on-outcome}, drawn from attacks independently known to succeed against the undefended agent (the host-security analogue of evaluating a control only against code already shown to execute), so reported bypass rates are conditional on attack potency: large \(N\) tightens Clopper--Pearson intervals~\cite{clopper1934} without buying generalization. Multi-token prediction (MTP) was treated as a per-engine decision (verify per-token grammar-mask correctness and run-to-run determinism, or disable it). In practice the per-trial MTP setting was not logged for three of the five result databases (\texttt{blast\_1p7b\_5m.db}, \texttt{mistral119\_anchor.db}, and \texttt{steelman\_abliterated.db}); the two closeout databases (\texttt{mistral119\_ladder.db} and \texttt{gemma31\_anchor.db}) stamp \texttt{engine\_commit} at runtime.\footnote{This is a self-disclosed reproducibility gap rather than a methodological claim; it is taken up in Section~\ref{sec:limits}.}
